---
title: "Probability and Bayesian Statistics"
author: "Prof. Sam Berchuck"
date: "2025-01-14"
date-format: "MMM DD, YYYY"
footer: "[🔗 BIOSTAT 725 - Spring 2025](https://biostat725-sp25.netlify.app/)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    html-math-method: mathjax
filters:
  - parse-latex
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r, echo=FALSE}
library(ggplot2)
library(gridExtra)
library(knitr)
```

# Agenda

-   Review of probability concepts

-   Introduction to Bayesian linear regression

-   Monte Carlo sampling

# Probability

This is foundational material that you should have already learned in a previous course. I'm reviewing important concepts that are needed for Bayesian inference.

## Review of probability

-   The goal of Bayesian statistics is to compute the posterior distribution (i.e., the uncertainty distribution of the parameters, $\boldsymbol{\theta}$, after observing the data, $\mathbf{Y}$)

-   This is the conditional distribution of $\boldsymbol{\theta}$ given $\mathbf{Y}$

-   Therefore, we need to review the probability concepts that lead to the conditional distribution of one variable conditioned on another

    1.  Probability mass (pmf) and density (pdf) functions

    2.  Joint distributions

    3.  Marginal and conditional distributions

## Random variables

-   $X$ (capital) is a random variable

-   We want to compute the probability that $X$ takes on a specific value $x$ (lowercase)

    -   This is denoted $P(X = x)$

-   We also might want to compute the probability of $X$ being in a set $\mathcal A$

    -   This is denoted $P(X \in \mathcal A)$

-   The set of possible values that $X$ can take on is called its support, $\mathcal S$

## Random variables - example

-   Example 1: $X$ is the roll of a die

    -   The support is $\mathcal S = \{1, 2, 3, 4, 5, 6\}$
    -   $P(X = 1) = 1/6$

-   Example 2: $X$ is a newborn baby’s weight

    -   The support is $\mathcal S = (0, \infty)$
    -   $P(X \in [0, \infty]) = 1$

<!-- ## Axioms of probability -->

<!-- For two events $A$ and $B$, -->

<!-- A1. Probabilities are between 0 and 1, importantly $P(A^c | A) = 0$ and $P(A | A) = 1$. -->

<!-- A2. If $A$ and $B$ are disjoint, then $P(A\text{ or }B) = P(A) + P(B)$. -->

<!-- A3. The joint probability of two events may be broken down stepwise: $P(A,B) = P(A | B)P(B)$. -->

<!-- ## Consequences of probability axioms -->

<!-- -   For any partition $\{H_i\}_{i = 1}^n$, $\sum_{i=1}^n P(H_i) = 1$ (rule of total probability) -->

<!--     -   Note: simplest partition $P(A) + P(A^c) = 1$ -->

<!-- -   $P(A) = \sum_{i=1}^n P(A, H_i)$ (rule of marginal probability) -->

<!--     -   Note: A3 implies that equivalently, $P(A) = \sum_{i=1}^n P(A | H_i) P(H_i)$ -->

<!-- -   $P(A | B) = P(A,B) / P(B)$ when $P(B) \neq 0$ -->

<!--     -   Note: these statements can also be made where each term is additionally conditioned on another event $C$ -->

## What is probability?

Objective (associated with frequentist)

-   $P(X = x)$ as a purely mathematical statement
-   If we repeatedly sampled $X$, the the proportion of draws equal to $x$ converges to $P(X = x)$

Subjective (associated with Bayesian)

-   $P(X = x)$ represents an individual’s degree of belief
-   Often quantified as the amount an individual would be willing to wager that $X$ will be $x$

A Bayesian analysis makes use of both of these concepts

## What is uncertainty?

Aleatoric uncertainty (likelihood)

-   Uncontrollable randomness in the experiment

-   For example, the results of a fair coin flip can never be predicted with certainty

Epistemic uncertainty (prior/posterior)

-   Uncertainty about a quantity that could theoretically be known
-   For example, if we flipped a coin infinitely-many times we could know the true probability of a head

A Bayesian analysis makes use of both of these concepts

## Probability versus statistics

Probability is the forward problem

-   We assume we know how the data are being generated and compute the probability of events
-   For example, what is the probability of flipping 5 straight heads if the coin is fair?

Statistics is the inverse problem

-   We use data to learn about the data-generating mechanism
-   For example, if we flipped five straight head, can we conclude the coin is biased?

Any statistical analysis obviously relies on probability

## Univariate distributions

-   We often distinguish between discrete and continuous random variables

-   The random variable $X$ is **discrete** if its support $\mathcal S$ is countable

-   Examples:

$X \in \{0, 1, 2, 3\}$ is the number of successes in 3 trials\

$X \in \{0, 1, 2, \ldots\}$ is the number of patients with COVID in Durham County

## Univariate distributions

-   We often distinguish between discrete and continuous random variables

-   The random variable $X$ is **continuous** if its support $\mathcal S$ is uncountable

-   Examples with $\mathcal S = (0, \infty)$:

$X > 0$ is systolic blood pressure\

$X > 0$ is a patient's BMI

## Discrete univariate distributions

-   If $X$ is discrete we describe its distribution with its **probability mass function (pmf)**

-   The pmf is $f(x) = P(X = x)$

-   The domain of $X$ is the set of $x$ with $f(x) > 0$

-   We must have $f(x) \geq 0$ and $\sum_x f(x) = 1$

-   The mean is $\mathbb E[X] = \sum_x x f(x)$

-   The variance is $\mathbb V(X) = \sum_x(x − \mathbb E[X])^2f(x)$

-   The last three sums are over $X$’s domain

## Parametric families of distributions

-   A statistical analysis typically proceeds by selecting a pmf that seems to match the distribution of a sample

-   We rarely know the pmf exactly, but we assume it is from a parametric family of distributions

-   For example, Binomial(10, 0.5) and Binomial(4, 0.1) are different but both from the binomial family

-   A family of distributions have the same equation for the pmf but differ by some unknown parameters $\boldsymbol{\theta}$

-   We must estimate these parameters

## Continuous univariate distributions

-   If $X$ is continuous we describe its distribution with the probability density function (pdf) $f(x) \geq 0$

-   Since there are uncountably many possible values, $P(X = x) = 0$ for all $x$,

-   Probabilities are computed as areas under the pdf curve $$P(a < X < b) = \int_a^b f(x)dx$$

-   Therefore, to be valid $f(x)$ must satisfy $f(x) \geq 0$ and $$P(−\infty < X < \infty) = \int_{-\infty}^{\infty} f(x)dx = 1$$

## Continuous univariate distributions

-   The domain is the set of $x$ values with $f(x) > 0$

-   The mean and the variance are defined similarly to the discrete case but with the sums replaced by integrals

-   The mean is $\mathbb E[X] = \int x f(x)dx$

-   The variance is $\mathbb V(X) = \int (x − \mathbb E[X])^2 f(x)dx$

## Joint distributions

-   $X = (X_1, \ldots, X_p)$ is a random vector (vectors and matrices should be in bold).

-   For notational convenience, let’s consider only $p = 2$ random variables $X$ and $Y$.

-   $(X, Y)$ is discrete if it can take on a countable number of values, such as\
    $X$ = number of hearts and $Y$ = number of face cards.

-   $(X, Y)$ is continuous if it can take on an uncountable number of values, such as\
    $X$ = birthweight and $Y$ = gestational age.

## Discrete random variables

-   The **joint pmf** is $f(x, y) = P(X = x, Y = y)$

    -   $\sum_x \sum_y f(x, y) = 1$

-   The **marginal pmf** for $X$ is $f_X(x) = P(X = x) = \sum_y f(x, y)$

-   The **marginal pmf** for $Y$ is $f_Y(y) = P(Y = y) = \sum_x
    f(x, y)$

-   The marginal distribution is the same as univariate distribution as if we ignored the other variable

## Discrete random variables

-   The **conditional pmf** of $Y$ given $X$ is $f(y|x) = P(Y = y|X = x) = \frac{P(X = x, Y = y)}{P(X = x)} = \frac{f(x, y)}{f_X (x)}.$

-   $X$ and $Y$ are **independent** if $f(x, y) = f_X(x)f_Y(y)$ for all $x$ and $y$

    -   Variables are dependent if they are not independent

-   Equivalently, $X$ and $Y$ are independent if $f(x|y) = f_X(x)$ for all $x$ and $y$

## Discrete random variables

-   Notation: $X_1, \dots, X_n \overset{\mathrm{iid}}{\sim} f(x)$ means that $X_1, \ldots, X_n$ are independent and identically distributed

-   This implies the **joint pmf** is $$P(X_1 = x_1, \ldots, X_n = x_n) = \prod_{i=1}^n f(x_i)$$

-   The same notation and definitions of independence apply to continuous random variables

-   In this class, assume independence unless otherwise noted

## Continuous random variables

-   Manipulating joint pdfs is similar to joint pmfs but sums are replaced by integrals

-   The **joint pdf** is denoted $f(x, y)$

-   Probabilities are computed as volume under the pdf: $$P((X, Y) ∈ A) = \int_A f(x, y)dxdy$$ where $A \subset \mathbb{R}^2$

## Continuous random variables

-   The **marginal pdf** of $X$ is $f_X(x) = \int f(x, y)dy$

-   $f_X$ is the univariate pdf for $X$ as if we never considered $Y$

-   The **conditional pdf** of $Y$ given $X$ is $$f(y|x) = \frac{f(x, y)}{f_X (x)}$$

-   Proper: $\int f(y|x)dy = \int \frac{f(x,y)}{f_X(x)}dy = \int \frac{f(x,y)dy}{f_X(x)} = 1$

## Defining joint distributions conditionally

-   Specifying joint distributions is hard

-   Every joint distribution can be written $f(x, y) = f(y|x)f(x)$

-   Therefore, any joint distribution can be defined by

    1.  $X$’s marginal distribution

    2.  The conditional distribution of $Y|X$

-   The joint problem reduces to two univariate problems

-   This idea forms the basis of hierarchical modeling

<!-- ## Review: Set theory {.small} -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- **set**: a collection of elements, denoted by {} -->

<!-- Examples -->

<!-- -   $\phi$ = {} "the empty set" -->

<!-- -   $A$ = {1, 2, 3} -->

<!-- -   $B$ = {taken BIOSTAT 724, has not taken BIOSTAT 724} -->

<!-- -   $C$ = {{1,2,3}, {4, 5}} -->

<!-- ::: -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- **subset**: denoted by $\subset$, $A \subset B$ iff -->

<!-- $a \in A \implies a \in B$ -->

<!-- Examples -->

<!-- Using the previously examples of $A$, $B$ and $C$ above, -->

<!-- -   $A \subset C$ -->

<!-- -   $A \not\subset B$ -->

<!-- ::: -->

<!-- Recall: $\cup$ means "union", "or"; $\cap$ means "intersection", "and" -->

<!-- ## Review: Set theory {.small} -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- **partition**: {$H_1, H_2, ... H_n$} = $\{H_i\}_{i = 1}^n$ is a -->

<!-- partition of $\mathcal{H}$ if -->

<!-- 1.  the union of sets is $\mathcal{H}$ i.e., -->

<!--     $\cup_{i = 1}^n H_i = \mathcal{H}$ -->

<!-- 2.  the sets are disjoint i.e., $H_i \cap H_j = \phi$ for all $i \neq j$ -->

<!-- ::: -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- **sample space**: $\mathcal{H}$, the set of all possible data sets -->

<!-- (outcomes) -->

<!-- **event**: a set of one or more outcomes -->

<!-- Note: p($\mathcal{H}$) = 1 -->

<!-- Examples -->

<!-- -   Roll a six-sided die once. The sample space -->

<!--     $\mathcal{H} = \{1, 2, 3, 4, 5, 6\}$. -->

<!-- -   Let $A$ be the event that the die lands on an even number. -->

<!--     $A = \{2, 4, 6 \}$ -->

<!-- ::: -->

## Bayes rule

Our course will focus on Bayes rule,

$$f(\boldsymbol{\theta}|\mathbf{Y}) = \frac{f(\mathbf{Y}|\boldsymbol{\theta})f(\boldsymbol{\theta})}{\int f(\mathbf{Y}|\boldsymbol{\theta})f(\boldsymbol{\theta})d\boldsymbol{\theta}}.$$

Bayes' rule tells us how to update beliefs about $\boldsymbol{\theta}$ given data $\mathbf{Y}$.

<!-- ## Independence -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- Two events $F$ and $G$ are **conditionally independent** given $H$ if -->

<!-- $p(F, G | H) = p(F | H) p(G | H)$ -->

<!-- This implies that $p(F | H, G) = p(F | H).$ -->

<!-- ::: -->

<!-- This means that if we know $H$, then $G$ does not supply any additional information about $F$. -->

<!-- ## Random variables -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- In Bayesian inference, a **random variable** is an unknown numerical quantity about which we make probability statements. -->

<!-- The **support** of a random variable is the set of values a random variable can take. -->

<!-- Examples, -->

<!-- -   Data. E.g., the amount of a wheat a field will yield later this year. Since this data has not yet been generated, the quantity is unknown. -->

<!-- -   A population parameter. E.g., the true mean resting heart rate of Duke students. Note: this is a *fixed* (non-random) quantity, but it is also *unknown*. We use probability to describe our uncertainty in this quantity. -->

<!-- ::: -->

<!-- ## Discrete random variables -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- **discrete** random variable: a random variable that takes countably -->

<!-- many values. $Y$ is discrete if its possible outcomes can be enumerated -->

<!-- $\mathcal{Y} = \{y_1, y_2, \ldots \}$. -->

<!-- Note: discrete does not mean finite. There may be infinitely many -->

<!-- outcomes! -->

<!-- Examples, -->

<!-- -   $Y$ = the number of children of a randomly sampled person -->

<!-- -   $Y$ = the number of visible stars in the sky on a randomly sampled -->

<!--     night of the year -->

<!-- For each $y \in \mathcal{Y}$, let $p(Y) = probability(Y = y)$. Then $p$ is the **probability mass function** (pmf) of the random variable $Y$. Finally, we have that, -->

<!-- $$ -->

<!-- \begin{aligned} -->

<!-- 0 \leq p(y) \leq 1,\\ -->

<!-- \sum_{y \in \mathcal{Y}} p(y) = 1. -->

<!-- \end{aligned} -->

<!-- $$ -->

<!-- ::: -->

<!-- ## Example discrete random variables {.small} -->

<!-- -   Binomial pmf: the probability of $y$ successes in $n$ trials, where -->

<!--     each trial has an individual probability of success $\theta$. -->

<!--     $$p(y | \theta) = {n \choose y} \theta ^y (1-\theta)^{n-y} \text{ for } y \in \{0, 1, \ldots n \}$$ -->

<!--     -   support: $y \in \{0, 1, 2, \ldots n\}$ -->

<!--     -   success probability $\theta \in [0, 1]$ -->

<!--     -   `dbinom(y, n, theta)` computes this pmf in R -->

<!-- -   Poisson pmf: probability of $y$ events occurring during a fixed -->

<!--     interval at a mean rate $\theta$ -->

<!--     $$p(y | \theta) = \frac{\theta^y e^{-\theta}}{y!}$$ -->

<!--     -   support: $y \in \{0, 1, 2, \ldots \}$ -->

<!--     -   rate $\theta \in \mathbb{R}^+$ -->

<!--     -   `dpois(y, theta)` computes this pmf in R -->

<!-- ## Continuous random variables -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- **continuous** random variable: a random variable that takes -->

<!-- *uncountably* many values. -->

<!-- The **probability density function** (pdf) of a continuous random -->

<!-- variable, $X$ is defined -->

<!-- $f(x) = pdf(x) = \lim_{\Delta x \rightarrow 0} \frac{p(x < X < x + \Delta x)}{\Delta x}$ -->

<!-- and the probability $X$ is in some interval, -->

<!-- $P(x_1 < X < x_2) = \int_{x_1}^{x_2} f(x) dx$.  -->

<!-- Finally, we have that, -->

<!-- $$ -->

<!-- \begin{aligned} -->

<!-- 0 \leq p(y) \ \text{and}, \\ -->

<!-- \int_{y \in \mathcal{Y}} p(y) = 1. -->

<!-- \end{aligned} -->

<!-- $$ -->

<!-- Note: For a *continuous random variable* $Y$, $f(y)$ can be larger than 1 -->

<!-- and $f(y)$ is **not** $P(Y = y)$, which equals 0. -->

<!-- ::: -->

<!-- ## Example continuous random variables {.small} -->

<!-- -   Normal pdf: $$ -->

<!--     f(x | \mu, \sigma) = (2\pi \sigma^2)^{-\frac{1}{2}}e^{-\frac{1}{2\sigma^2}(x-\mu)^2} -->

<!--     $$ -->

<!-- -   Uniform pdf: $$f(x|a,b) =  -->

<!--     \begin{cases} -->

<!--     \frac{1}{b - a} \hspace{.6cm}\text{ for } x \in [a, b]\\ -->

<!--     0 \hspace{1cm}\text{ otherwise } -->

<!--     \end{cases}$$ -->

<!-- ## Kernel -->

<!-- ::: callout-note -->

<!-- ## Definition -->

<!-- The part of the density/mass function that depends on the variable is -->

<!-- called the **kernel**. -->

<!-- Example: -->

<!-- -   the kernel of the normal pdf is $e^{-\frac{(x-\mu)^2}{2\sigma^2}}$ -->

<!-- ::: -->

<!-- ## Moments {.smaller} -->

<!-- For a random variable $X$, the $n$th moment is defined as $\mathbb{E}[X^n]$. -->

<!-- Recall, the expected value is defined for discrete random variable $X$,  -->

<!-- $$ -->

<!-- \mathbb{E}[X] = \sum_{x \in \mathcal{X}} x p(x) -->

<!-- $$  -->

<!-- and for continuous random variable $Y$, -->

<!-- $$ -->

<!-- \mathbb{E}[Y] = \int_{-\infty}^{\infty} y f(y) dy -->

<!-- $$  -->

<!-- The variance of a random variable, is also known as the *second -->

<!-- central moment* and is defined -->

<!-- $$ -->

<!-- \mathbb{E}[(X - \mathbb{E}[X])^2] -->

<!-- $$  -->

<!-- or equivalently, $\mathbb{E}[X^2] - \mathbb{E}[X]^2$. More generally, the covariance between two random variables $X$ and $Y$ -->

<!-- is defined as, -->

<!-- $$ -->

<!-- \mathbb{E}[(X - \mathbb{E}[X])(Y - \mathbb{E}[Y])]. -->

<!-- $$ -->

# Bayesian Linear Regression: Closed Form Posterior

## Defining the model

Suppose we have an observation $Y_i$ for subject $i$ ($i=1,\ldots,n$), that is modeled as follows,

\begin{align*}
Y_i &= \beta_0 + x_{i1} \beta_1 + \cdots + x_{ip} \beta_p + \epsilon_i, \quad \epsilon_i \sim N(0, \sigma^2)\\
&= \mathbf{x}_i \boldsymbol{\beta} + \epsilon_i.
\end{align*}

-   $\mathbf{x}_i = (1, x_{i1},\ldots,x_{ip})$ is a $(p+1)$-dimensional row vector of covariates (and intercept).
-   $\boldsymbol{\beta} = (\beta_0, \beta_1,\ldots,\beta_p)^\top$ is a $(p+1)$-dimensional column vector of population regression parameters.
-   $\epsilon_i$ is a Gaussian measurement error term with variance $\sigma^2$.
-   For the purpose of today's lecture, we assume $\sigma$ is known.

## Defining the likelihood

The individual likelihood contribution for subject $i$ is given by, $$Y_i|\boldsymbol{\beta} \sim N(\mathbf{x}_i \boldsymbol{\beta}, \sigma^2) \Leftrightarrow f(Y_i|\boldsymbol{\beta}) = N(\mathbf{x}_i \boldsymbol{\beta}, \sigma^2),$$ and the full data likelihood (or observed data likelihood) is given by, $$f(\mathbf{Y} | \boldsymbol{\beta}) = \prod_{i=1}^n f(Y_i|\boldsymbol{\beta}),$$ where $\mathbf{Y} = (Y_1,\ldots,Y_n)$.

## Defining the likelihood (matrix version)

We can also write the likelihood directly, $$\mathbf{Y} | \boldsymbol{\beta} \sim N(\mathbf{X} \boldsymbol{\beta}, \sigma^2 \mathbf{I}_n),$$ where $\mathbf{X}$ is an $n \times (p + 1)$ dimensional matrix with row $\mathbf{x}_i$ and $\mathbf{I}_n$ is an $n$-dimensional identity matrix. Thus, the mean of the observed data is modeled as a linear function of the parameters,

$$
\mathbb{E}[ \mathbf{Y} | \boldsymbol{\beta} ]  =
  \begin{bmatrix}
    1 & x_{12} & \ldots & x_{1p} \\
    1 & x_{22} & \ldots & x_{2p} \\
    \vdots & \vdots & & \vdots\\
    1 & x_{n2} & \ldots & x_{np}
  \end{bmatrix}
    \begin{bmatrix}
    \beta_0\\
    \beta_1\\
    \vdots\\
    \beta_p
  \end{bmatrix} = \mathbf{X} \boldsymbol{\beta}.
$$

## Linear regression estimation

-   Ordinary least squares (OLS)

<center>$\hat{\boldsymbol{\beta}}_{\text{OLS}} = \underset{\boldsymbol{\beta}}{\operatorname{argmin}} || \mathbf{Y} - \mathbf{X}\boldsymbol{\beta}||^2$</center>

-   Maximum likelihood estimation (MLE)

<center>$\hat{\boldsymbol{\beta}}_{\text{MLE}} = \underset{\boldsymbol{\beta}}{\operatorname{argmax}} f(\mathbf{Y} | \boldsymbol{\beta})$</center>

-   $\hat{\boldsymbol{\beta}}_{\text{OLS}} = \hat{\boldsymbol{\beta}}_{\text{MLE}} = (\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y}$

## Bayesian estimation

Estimation in a Bayesian setting is uniquely encoded through the posterior, \begin{align*}
f(\boldsymbol{\beta} | \mathbf{Y}) &= \frac{f(\mathbf{Y}, \boldsymbol{\beta})}{f(\mathbf{Y})}\\
&= \frac{f(\mathbf{Y} | \boldsymbol{\beta}) f(\boldsymbol{\beta})}{f(\mathbf{Y})}\\
&= \frac{f(\mathbf{Y} | \boldsymbol{\beta}) f(\boldsymbol{\beta})}{\int f(\mathbf{Y} | \boldsymbol{\beta}) f(\boldsymbol{\beta})d\boldsymbol{\beta}}.
\end{align*}

To compute the posterior distribution, we need the prior $f(\boldsymbol{\beta})$.

## Prior definition

Let's assume that the prior for $\boldsymbol{\beta}$ is Gassian,

$$f(\boldsymbol{\beta}) = N(\boldsymbol{\beta}_0,\sigma_{\beta}^2 \mathbf{I}_{p+1}).$$

-   $\boldsymbol{\beta}_0$ is the prior mean (i.e., our a-priori guess for the likely value of $\boldsymbol{\beta}$)
-   $\sigma_{\beta}^2$ is the prior variance (i.e., encodes our certainty for our a-priori guess)

## Computing the posterior

With this prior, the posterior can be found in closed-form. The posterior is Gaussian with the following moments: \begin{align*}
\mathbb{V}(\boldsymbol{\beta} | \mathbf{Y}) &= \left(\frac{\mathbf{I}_{p+1}}{\sigma_{\beta}^2} + \frac{\mathbf{X}^\top \mathbf{X}}{\sigma^2}\right)^{-1}\\
\mathbb{E}[\boldsymbol{\beta} | \mathbf{Y}] &= \left(\frac{\mathbf{I}_{p+1}}{\sigma_{\beta}^2} + \frac{\mathbf{X}^\top \mathbf{X}}{\sigma^2}\right)^{-1}\left(\frac{\boldsymbol{\beta}_0}{\sigma_{\beta}^2} + \frac{\mathbf{X}^\top \mathbf{Y}}{\sigma^2}\right)
\end{align*}

## Techniques to find this posterior

-   Brute force: complete the square

-   Easy: kernel recognition

::: callout-note
## Definition

The part of the pdf/pmf that depends on the variable is called the **kernel**.

Example:

-   The kernel of the multivariate normal pdf, $\mathbf{Y} \sim N(\boldsymbol{\mu}, \boldsymbol{\Sigma})$, is

\begin{align*}
f(\mathbf{Y} | \boldsymbol{\mu}, \boldsymbol{\Sigma}) &\propto \exp\left\{-\frac{1}{2}\left[\left(\mathbf{Y} - \boldsymbol{\mu}\right)^\top \boldsymbol{\Sigma}^{-1}\left(\mathbf{Y} - \boldsymbol{\mu}\right)\right]\right\}\\
&\propto \exp\left\{-\frac{1}{2}\left[\mathbf{Y}^\top \boldsymbol{\Sigma}^{-1}\mathbf{Y} - 2\mathbf{Y}^\top \boldsymbol{\Sigma}^{-1} \boldsymbol{\mu}\right]\right\}
\end{align*}
:::

## Using the kernel to find the posterior {.small}

\begin{align*}
f(\boldsymbol{\beta} | \mathbf{Y}) &\propto f(\mathbf{Y} | \boldsymbol{\beta}) f(\boldsymbol{\beta})\\
&\propto \exp\left\{-\frac{1}{2}\left[\frac{\left(\mathbf{Y} - \mathbf{X}\boldsymbol{\beta}\right)^\top\left(\mathbf{Y} - \mathbf{X}\boldsymbol{\beta}\right)}{\sigma^2}\right]\right\}\\
&\quad \times \exp\left\{-\frac{1}{2}\left[\frac{\left(\boldsymbol{\beta} - \boldsymbol{\beta}_0\right)^\top\left(\boldsymbol{\beta}_0 - \boldsymbol{\beta}_0\right)}{\sigma_{\beta}^2}\right]\right\}\\
&\propto \exp\left\{-\frac{1}{2}\left[\boldsymbol{\beta}\left(\frac{\mathbf{X}^\top \mathbf{X}}{\sigma^2} + \frac{\mathbf{I}_n}{\sigma_{\beta}^2} \right)\boldsymbol{\beta} - 2\boldsymbol{\beta}^\top\left(\frac{\mathbf{X}^\top\mathbf{Y}}{\sigma^2} + \frac{\boldsymbol{\beta}_0}{\sigma_{\beta}^2}\right)\right]\right\}\\
&\propto \exp\left\{-\frac{1}{2}\left[\boldsymbol{\beta}\mathbf{A}\boldsymbol{\beta} - 2\boldsymbol{\beta}^\top\mathbf{a}\right]\right\}
\end{align*}

This is the kernel of a multivariate normal for $\boldsymbol{\beta}$, with $\mathbf{A} = \boldsymbol{\Sigma}^{-1}$ and $\mathbf{a} = \boldsymbol{\Sigma}^{-1}\boldsymbol{\mu}$. It's easy to see then that, $f(\boldsymbol{\beta} | \mathbf{Y}) = N(\mathbf{A}^{-1}\mathbf{a},\mathbf{A}^{-1}).$

::: callout-note
## Recall the kernel for the multivariate normal: $\exp\left\{-\frac{1}{2}\left[\mathbf{Y}^\top \boldsymbol{\Sigma}^{-1}\mathbf{Y} - 2\mathbf{Y}^\top \boldsymbol{\Sigma}^{-1} \boldsymbol{\mu}\right]\right\}$
:::

## Back to the posterior

With this prior, the posterior can be found in closed-form. The posterior is Gaussian with the following moments: \begin{align*}
\mathbb{V}(\boldsymbol{\beta} | \mathbf{Y}) &= \left(\frac{\mathbf{I}_{p+1}}{\sigma_{\beta}^2} + \frac{\mathbf{X}^\top \mathbf{X}}{\sigma^2}\right)^{-1}\\
\mathbb{E}[\boldsymbol{\beta} | \mathbf{Y}] &= \left(\frac{\mathbf{I}_{p+1}}{\sigma_{\beta}^2} + \frac{\mathbf{X}^\top \mathbf{X}}{\sigma^2}\right)^{-1}\left(\frac{\boldsymbol{\beta}_0}{\sigma_{\beta}^2} + \frac{\mathbf{X}^\top \mathbf{Y}}{\sigma^2}\right)
\end{align*}

. . .

-   $\sigma_{\beta}^2 \rightarrow \infty: \mathbb{E}[\boldsymbol{\beta} | \mathbf{Y}] = (\mathbf{X}^\top \mathbf{X})^{-1}\mathbf{X}^\top \mathbf{Y} = \hat{\boldsymbol{\beta}}_{\text{OLS}} = \hat{\boldsymbol{\beta}}_{\text{MLE}}$

. . .

-   $\sigma^2 \rightarrow \infty: \mathbb{E}[\boldsymbol{\beta} | \mathbf{Y}] = \boldsymbol{\beta}_0$

## How can we use the posterior?

Let's simulate some data:

```{r}
###True parameters
sigma <- 1.5 # true measurement error
beta <- matrix(c(-1.5, 3), ncol = 1) # true beta

###Simulation settings
n <- 100 # number of observations
p <- length(beta) - 1 # number of covariates

###Simulate data
set.seed(54) # set seed
X <- cbind(1, matrix(rnorm(n * p), ncol = p))
Y <- as.numeric(X %*% beta + rnorm(n, 0, sigma))
```

## Visualize simulated data

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-asp: 1
#| layout-ncol: 2
dat.fig <- data.frame(x = X[, 2], y = Y)
ggplot(dat.fig, aes(x = x, y = y)) + 
  geom_point()
ggplot(dat.fig, aes(x = y)) +
  geom_histogram()
```

## Inspecting the prior

First, we define hyperparameteters.

```{r}
beta0 <- matrix(0, nrow = p + 1, ncol = 1)
sigma_beta <- 10
```

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4
#| fig-asp: 1
#| layout-ncol: 3
dat.fig <- data.frame(beta0 = rnorm(1000, beta0[1], sigma_beta), 
           beta1 = rnorm(1000, beta0[1], sigma_beta))
ggplot(dat.fig, aes(x = beta0, y = beta1)) + 
  geom_point() + 
  xlab(expression(beta[0])) + 
  ylab(expression(beta[1]))
ggplot(dat.fig, aes(x = beta0)) + 
  geom_density() + 
  xlab(expression(beta[0])) + 
  ylab("Density")
ggplot(dat.fig, aes(x = beta1)) + 
  geom_density() + 
  xlab(expression(beta[1])) + 
  ylab("Density")
```

## Comparison with OLS/MLE

Compute posterior moments

```{r}
var_beta <- chol2inv(chol(t(X) %*% X / sigma^2 + diag(p + 1) / sigma_beta^2))
mean_beta <- var_beta %*% (beta0 / sigma_beta^2 + t(X) %*% Y / sigma^2)
```

```{r}
#| echo: false
out <- data.frame(parameter = c("intercept", "slope"),
  true = beta, 
  bayes = mean_beta, 
  ols = coef(lm(Y ~ -1 + X)))
colnames(out)[4] <- "ols/mle"
rownames(out) <- NULL
kable(out, format = "html") 
```

<!-- ## Summarizing the posterior -->

<!-- - Summarizing a posterior (with known parametric form) is straightforward for basic quantities of interest -->

<!-- 1. Posterior means, medians, modes, and variances -->

<!-- 2. Joint, conditional, and marginal probabilities, for example:  $P(\beta_j < c | \mathbf{Y})$ -->

<!-- 3. $\alpha$-quantiles: $\{q_{\alpha} : P(\beta_j < q_{\alpha} | \mathbf{Y}) = \alpha\}, \alpha \in (0,1)$ -->

<!-- 4. $\ldots$ -->

## Summarizing the posterior

-   Posteriors are often not available in closed form.

-   Even when we have a closed form posterior, it can be difficult to compute summaries of interest.

-   For example, consider $P(\beta_j < c |\mathbf{Y})$. What are our options to calculate this probability?

    -   Direct integration (by hand)

    -   Numerical integration/software packages (e.g., $\texttt{pnorm}$)

These methods work well for standard posterior quantities and distributions

## Summarizing the posterior

However, sometimes we will want to summarize other aspects of a posterior distribution

-   $P(\beta_j \in A|\mathbf{Y})$ for some arbitrary set $A$

-   Means and standard deviations of some function of $\beta_j$, $g\left(\beta_j\right)$

-   The posterior distribution of functions of many parameters:

    -   $|\beta_1 - \beta_2|$, $\beta_1/\beta_2$, $\max\left\{\beta_1,\ldots,\beta_p\right\}$, $\dots$

-   Obtaining exact values for these posterior quantities can be difficult or even impossible

What are our options?

# Introduction to Monte Carlo approximation

## Monte Carlo (MC) approximation

-   Integration method based on random sampling

-   The general principles and procedures remain relatively constant across a broad class of problems

-   Suppose we have $S$ iid samples from our posterior distribution: $\boldsymbol{\beta}^{\left(1\right)},\ldots,\boldsymbol{\beta}^{\left(S\right)}\stackrel{\text{iid}}{\sim}f\left(\boldsymbol{\beta}|\mathbf{Y}\right)$

-   Then the empirical distribution of $\left\{\boldsymbol{\beta}^{\left(1\right)},\ldots,\boldsymbol{\beta}^{\left(S\right)}\right\}$ would approximate $f\left(\boldsymbol{\beta}|\mathbf{Y}\right)$, with the approximation improving as $S$ increases

-   The empirical distribution of $\left\{\boldsymbol{\beta}^{\left(1\right)},\ldots,\boldsymbol{\beta}^{\left(S\right)}\right\}$ is known as a MC approximation to $f\left(\boldsymbol{\beta}|\mathbf{Y}\right)$

## MC approximation

Let $g\left(\boldsymbol{\beta}\right)$ be (just about) any function of $\boldsymbol{\beta}$. The law of large numbers says that if,

$$\boldsymbol{\beta}^{\left(1\right)},\ldots,\boldsymbol{\beta}^{\left(S\right)}\stackrel{\text{iid}}{\sim}f\left(\boldsymbol{\beta}|\mathbf{Y}\right),$$ then, $$\frac{1}{S}\sum_{s=1}^S g\left(\boldsymbol{\beta}^{\left(s\right)}\right)\rightarrow \mathbb{E}\left[g\left(\boldsymbol{\beta}\right)|\mathbf{Y}\right]=\int g\left(\boldsymbol{\beta}\right)f\left(\boldsymbol{\beta}|\mathbf{Y}\right)d\boldsymbol{\beta},$$ as $S\rightarrow \infty$.

## MC approximation

-   Implications (as $S\rightarrow \infty$):

    -   $\overline{\boldsymbol{\beta}}=\frac{1}{S}\sum_{s=1}^S \boldsymbol{\beta}^{\left(s\right)} \rightarrow \mathbb{E}\left[\boldsymbol{\beta}|\mathbf{Y}\right]$

    -   $\frac{1}{S-1}\sum_{s=1}^S \left(\boldsymbol{\beta}^{\left(s\right)}-\overline{\boldsymbol{\beta}}\right)^2 \rightarrow \mathbb{V}\left(\boldsymbol{\beta}|\mathbf{Y}\right)$

    -   $\frac{1}{S}\sum_{s=1}^S 1\left(\boldsymbol{\beta}^{\left(s\right)}\leq \mathbf{c}\right) \rightarrow P\left(\boldsymbol{\beta}\leq \mathbf{c} | \mathbf{Y}\right)$

    -   $\alpha$-quantile of $\left\{\boldsymbol{\beta}^{\left(1\right)},\ldots,\boldsymbol{\beta}^{\left(S\right)}\right\}\rightarrow q_{\alpha}$

-   Just about any aspect of the posterior distribution can be approximated arbitrarily exactly with a large enough MC sample

## Posterior inference for arbitrary functions

Interest in the posterior distribution of a function of $\boldsymbol{\beta}$, $g\left(\boldsymbol{\beta}\right)$

-   MC sampling plan:

    -   Draw $\boldsymbol{\beta}^{\left(1\right)}$ from $f\left(\boldsymbol{\beta}|\mathbf{Y}\right)$, calculate $g\left(\boldsymbol{\beta}^{\left(1\right)}\right)$

        $\ldots$

    -   Draw $\boldsymbol{\beta}^{\left(S\right)}$ from $f\left(\boldsymbol{\beta}|\mathbf{Y}\right)$, calculate $g\left(\boldsymbol{\beta}^{\left(S\right)}\right)$

    $\Rightarrow g\left(\boldsymbol{\beta}^{\left(1\right)}\right),\ldots,g\left(\boldsymbol{\beta}^{\left(S\right)}\right)\stackrel{\text{iid}}{\sim}f\left(g\left(\boldsymbol{\beta}\right)|\mathbf{Y}\right)$

-   Therefore, similar quantities can be estimated (posterior mean, variance, quantiles, distribution, etc.)

## How many samples to take? {.small}

We can use a central limit theorem: $\sqrt{S}\left(\overline{\boldsymbol{\beta}}-\mathbb{E}[\boldsymbol{\beta} | \mathbf{Y}]\right)/\sigma \stackrel{d}{\rightarrow} \text{N}\left(0,1\right)$,

where $\sigma^2 = \mathbb{V}\left(\overline{\boldsymbol{\beta}}\right) = \frac{1}{S^2}\sum_{s=1}^S\mathbb{V}\left(\boldsymbol{\beta}^{(s)}\right) = \frac{1}{S}\mathbb{V}\left(\boldsymbol{\beta} | \mathbf{Y}\right)$.

$\implies \overline{\boldsymbol{\beta}}\approx N\left(\mathbb{E}[\boldsymbol{\beta} | \mathbf{Y}],\sigma^2/S\right)$

-   MC standard error: $\sqrt{\frac{\widehat{\sigma}^2}{S}}$, $\widehat{\sigma}^2=\frac{1}{S-1}\sum_{s=1}^S\left(\boldsymbol{\beta}^{\left(s\right)}-\overline{\boldsymbol{\beta}}\right)^2$

-   Approximate 95% MC confidence interval for the posterior mean: $\overline{\boldsymbol{\beta}} \pm 2\sqrt{\frac{\widehat{\sigma}^2}{S}}$

-   Choose $S$ large enough to report the posterior mean with your desired precision

-   Reporting MC standard errors for the posterior mean is a good way to indicate that $S$ is large enough

## Returning to linear regression

Let's obtain $S = 1,000$ samples from our posterior.

```{r}
library(mvtnorm) # multivariate rng
beta_samples <- rmvnorm(1000, mean_beta, var_beta)
```

We can compute the posterior mean and variance.

```{r}
print(apply(beta_samples, 2, mean))
print(apply(beta_samples, 2, var))
```

## Assessing accuracy

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 10
#| layout-ncol: 1
out <- matrix(nrow = 2, ncol = 5000)
for (s in 1:5000) {
  beta_samples <- rmvnorm(s, mean_beta, var_beta)
  out[, s] <- apply(beta_samples, 2, mean)
}
dat.fig <- data.frame(S = rep(1:5000, 2),
                      mean = as.numeric(t(out)),
                      parameter = rep(c("beta0", "beta1"), each = 5000))
true_beta <- data.frame(true = mean_beta,
                        parameter = c("beta0", "beta1"))
ggplot(dat.fig, aes(x = S, y = mean)) + 
  geom_line() + 
  geom_hline(data = true_beta, aes(yintercept = true), color = "red", lwd = 1.5) +
  facet_grid(parameter ~ ., scales = "free") + 
  ylab("Posterior mean estimate")
```

## Additional posterior summaries

```{r}
# median
apply(beta_samples, 2, median)
# 95% credible intervals
apply(beta_samples, 2, function(x) quantile(x, probs = c(0.025, 0.975)))
# evaluating probability
mean(beta_samples[, 1] < -1.5)
# summarizing arbitrary functions of the parameters
beta_new <- beta_samples[, 1] * beta_samples[, 2]^3
c(mean(beta_new), quantile(beta_new, probs = c(0.025, 0.975)))
```
