---
title: "Model Comparison"
author: "Prof. Sam Berchuck"
date: "2026-01-29"
date-format: "MMM DD, YYYY"
footer: "[🔗 BIOSTAT 725 - Spring 2026](https://biostat725-sp26.netlify.app/)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    html-math-method: mathjax
filters:
  - parse-latex
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(knitr)
library(mvtnorm)
library(coda)
```

## Review of last lecture

On Tuesday, we learned about various ways to check MCMC convergence and model fit.

-   Traceplots, effective sample size ($n_{eff}$), MC standard error, $\hat{R}$, sampling issues

-   Posterior predictive checks

-   Model checks using `shinystan`

Today, we will learn about model comparisons.

## Model comparison

-   In statistical modeling, a more complex model almost always results in a better fit to the data.

    -   A more complex model means one with more parameters.

-   If one has 10 observations, one can have a model with 10 parameters that can perfectly predict every single data point (by just having a parameter to predict each data point).

-   There are two problems with overly complex models.

    1.  They become increasingly hard to interpret (think a straight line versus a polynomial).

    2.  They are more at risk of overfitting, such that it does not work for future observations.

## Model fit: an example data set

-   Let's explore the idea of model fit using an example dataset from the `openintro` package called `bdims`.

-   This dataset contains body girth measurements and skeletal diameter measurements.

-   Today we will explore the association between height and weight.

## Model fit: an example data set

```{r}
library(openintro)
dat <- data.frame(weight = bdims$wgt * 2.20462, # convert weight to lbs
                  height = bdims$hgt * 0.393701, # convert height to inches
                  sex = ifelse(bdims$sex == 1, "Male", "Female"))
```

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 6
#| fig-height: 4
#| layout-ncol: 1
set.seed(1)
dat_sub <- dat[sample(1:nrow(dat), 10), ]
ggplot(dat, aes(x = height, y = weight)) + 
  geom_point() + 
  geom_point(data = dat_sub, aes(x = height, y = weight), color = "red") +
  labs(x = "Height (inches)", y = "Weight (pounds)")
```

## Models of increasing complexity {.midi}

-   When using height to predict weight, we can use models of increasing complexity using higher order polynomials.

-   Let's fit the following models to the subset of 10 data points:

\begin{align*}
\mathbb{E}[weight_i] &= \beta_0 + \beta_1 height_i\\
\mathbb{E}[weight_i] &= \beta_0 + \beta_1 height_i + \beta_2 height_i^2\\
\mathbb{E}[weight_i] &= \beta_0 + \beta_1 height_i + \beta_2 height_i^2 + \beta_3 height_i^3\\
\mathbb{E}[weight_i] &= \beta_0 + \beta_1 height_i + \beta_2 height_i^2 + \beta_3 height_i^3 + \beta_4 height_i^4
\end{align*}

-   We can compare these models using standard measures of goodness-of-fit, including $R^2$ and root mean squared error (RMSE).

## Overfitting and underfitting

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.25
#| fig-height: 2.5
#| layout-ncol: 2
reg1 <- lm(weight ~ height, data = dat_sub)
reg2 <- lm(weight ~ height + I(height^2), data = dat_sub)
reg3 <- lm(weight ~ height + I(height^2) + I(height^3), data = dat_sub)
reg4 <- lm(weight ~ height + I(height^2) + I(height^3) + I(height^4), data = dat_sub)
reg5 <- lm(weight ~ height + I(height^2) + I(height^3) + I(height^4) + I(height^5), data = dat_sub)
reg6 <- lm(weight ~ height + I(height^2) + I(height^3) + I(height^4) + I(height^5) + I(height^6), data = dat_sub)
regs <- list(reg1, reg2, reg3, reg4, reg5, reg6)
round2 <- function(x) format(round(x, 2), nsmall = 2)
labels <- c("Linear", "Quadratic", "Cubic", "4th degree polynomial", "5th degree polynomial", "6th degree polynomial")
for (i in 1:4) {
  reg <- regs[[i]]
  r2 <- summary(reg)$r.squared
  rmse <- sqrt(mean((predict(reg) - dat_sub$weight)^2))
  x <- seq(min(dat$height), max(dat$height), length.out = 1001)
  dat_curve <- data.frame(x = x,
                          y = predict(reg, newdata = data.frame(height = x)))
  p <- ggplot(dat_sub, aes(x = height, y = weight)) + 
    geom_point(color = "red") + 
    # geom_line(data = dat_curve, aes(x = x, y = y), color = "blue", lwd = 1.5) + 
    geom_smooth(method = "lm", formula = y ~ poly(x, i), se = TRUE, fullrange = TRUE, lwd = 1.5) +
    ylim(min(dat$weight), max(dat$weight)) + 
    xlim(min(dat$height), max(dat$height)) + 
    labs(x = "Height (inches)", y = "Weight (pounds)", subtitle = labels[i]) + 
    annotate("text", x = 58, y = 250, label = as.expression(bquote(R^2 == ~ .(round2(r2)))), hjust = 0, vjust = 0) + 
    annotate("text", x = 58, y = 235, label = as.expression(bquote("RMSE" == ~ .(round2(rmse)))), hjust = 0, vjust = 0)
  print(p)
}
```

## Overfitting and underfitting

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 4.25
#| fig-height: 2.5
#| layout-ncol: 2
for (i in 1:4) {
  reg <- regs[[i]]
  rmse <- sqrt(mean((predict(reg, newdata = data.frame(height = dat$height)) - dat$weight)^2))
  x <- seq(min(dat$height), max(dat$height), length.out = 1001)
  dat_curve <- data.frame(x = x,
                          y = predict(reg, newdata = data.frame(height = x)))
  p <- ggplot(dat_sub, aes(x = height, y = weight)) + 
    geom_point(data = dat, aes(x = height, y = weight)) + 
    geom_point(color = "red") + 
    geom_line(data = dat_curve, aes(x = x, y = y), color = "blue", lwd = 1.5) + 
    # geom_smooth(method = "lm", formula = y ~ poly(x, i), se = FALSE, full_range = TRUE) + 
    ylim(min(dat$weight), max(dat$weight)) + 
    labs(x = "Height (inches)", y = "Weight (pounds)", subtitle = labels[i]) + 
    annotate("text", x = 72, y = 100, label = as.expression(bquote("RMSE" == ~ .(round2(rmse)))), hjust = 0, vjust = 0)
  print(p)
}
```

## Overfitting and underfitting

-   With more complex models, out-of-sample prediction becomes worse.

-   This is because when you use a complex model in a data set, it tailors the coefficients to any sampling errors and noise in the data such that it will not generalize to new observations.

-   Therefore, our goal in model comparison is to choose a model with the following two properties:

    1.  It is complex enough to capture the essence of the data generation process (and thus avoid underfitting),

    2.  It avoids overfitting to make the model usefull for predicting new observations.

## Finding an optimal model {.midi}

-   Trade-off between overfitting and underfitting (in machine learning this is commonly called bias-variance trade-off).

    -   A simple model tends to produce biased predictions because it does not capture the essence of the data generating process.

    -   A model that is overly complex is unbiased but results in a lot of uncertainty in the prediction.

-   Polynomials are merely one example of comparing simple to complex models. You can think about:

    -   Models with and without interactions,

    -   Models with a few predictors versus hundreds of predictors,

    -   Regression analyses versus hierarchical models, etc.

## Model Comparison {.midi}

-   When comparing models, we prefer models that are closer to the *true* data-generating process.

-   We need some ways to quantify the degree of *closeness* to the true model. Note that in this context models refer to the distributional family as well as the parameter values.

-   For example, the model $Y_i \sim N(5,2)$ is a different model than $Y_i \sim N(3,2)$, which is a different model than $Y_i \sim Gamma(2,2)$.

    -   The first two have the same family but different parameter values (different means, same SD), whereas the last two have different distributional families (Normal vs. Gamma).

-   One way to quantify the degree of *closeness* to the true model is using Kullback-Leibler (KL) divergence.

## Kullback-Leibler divergence {.midi}

-   For two models, $M_0$ and $M_1$, the KL divergence is given by,

\begin{align*}
D_{KL}\left(M_0 | M_1\right) &= \int_{-\infty}^{\infty} f_{M_0}(\mathbf{Y}) \log\frac{f_{M_0}(\mathbf{Y})}{f_{M_1}(\mathbf{Y})} d\mathbf{Y}\\
&\hspace{-1.5in}= \int_{-\infty}^{\infty} f_{M_0}(\mathbf{Y}) \log f_{M_0}(\mathbf{Y})d\mathbf{Y} - \int_{-\infty}^{\infty} f_{M_0}(\mathbf{Y}) \log f_{M_1}(\mathbf{Y})d\mathbf{Y}
\end{align*}

-   Note that $D_{KL}$ is not considered a distance, because it is not strictly symmetric, $D_{KL}\left(M_0 | M_1\right) \neq D_{KL}\left(M_1 | M_0\right)$.

## Kullback-Leibler divergence {.midi}

As an example, assume that the data are generated by a true model $M_0$, and we have two candidate models $M_1$ and $M_2$, where

::::: columns
::: {.column width="40%"}
-   $M_0: Y_i \sim N(3,2)$

-   $M_1: Y_i \sim N(3.5, 2.5)$

-   $M_2: Y_i \sim Cauchy(3,2)$
:::

::: {.column width="60%"}
```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 3
#| layout-ncol: 1
x <- seq(-4, 12, length.out = 1001)
dens1 <- dnorm(x, 3, 2)
dens2 <- dnorm(x, 3.5, 2.5)
dens3 <- dcauchy(x, 3, 2)
dat.fig <- data.frame(
  x = rep(x, 3),
  y = c(dens1, dens2, dens3),
  Model = rep(c("M[0]", "M[1]", "M[2]"), each = 1001)
)
ggplot(dat.fig, aes(x = x, y = y, color = Model)) + 
  geom_line(lwd = 1.5) + 
  labs(x = "Y", y = "Density") +
  scale_colour_manual(name = "Model", values = c("blue", "red", "green"), 
                      labels = c(expression(M[0]), expression(M[1]), expression(M[2]))) 
```
:::
:::::

-   $D_{KL}(M_0 |M_1) = 0.063$, $D_{KL}(M_0 | M_2) = 0.259$, so $M_1$ is a better model than $M_2$.

## Comparing models using KL {.midi}

-   Note that in the expression of $D_{KL}$, when talking about the same target model, the first term is always the same and describes the *true* model, $M_0$.

-   Therefore, it is sufficient to compare models on the second term,\
    $$\int_{-\infty}^{\infty} f_{M_0}(\mathbf{Y}) \log f_{M_1}(\mathbf{Y})d\mathbf{Y},$$ which can also be written as, $\mathbb{E}_{M_0}\left[\log f_{M_1}(\mathbf{Y})\right].$

-   This term is the **expected log predictive density (elpd)**.

-   A larger elpd is preferred. Why?

## Comparing models using KL

-   In the real world, we do not know $M_0$.

    -   If we knew, then we would just need to choose $M_0$ as our model and there will be no problem about model comparisons.

    -   Even if we knew the true model, we would still need to estimate the parameter values.

-   Thus, we cannot compute elpd, since the expectation is over $f_{M_0}(\mathbf{Y})$.

-   We need to estimate elpd!

## Comparing models using KL {.midi}

-   elpd is an expectation, so we can think about estimating it using Monte Carlo sampling, $$\frac{1}{S}\sum_{s = 1}^S\log f_{M_1}\left(\mathbf{Y}^{(s)}\right)\rightarrow \mathbb{E}_{M_0}\left[\log f_{M_1}(\mathbf{Y})\right], \quad \mathbf{Y}^{(s)} \sim f_{M_0}(\mathbf{Y}).$$

    -   We need to find a way to approximate, $f_{M_0}\left(\mathbf{Y}^{(s)}\right)$.

-   A naive way to approximate $f(\mathbf{Y}^{(s)})$ is to assume that the distribution of the observed data is the true model.

    -   This is equivalent to assuming that $\mathbf{Y}^{(s)} \sim \{\mathbf{Y}_1,\ldots,\mathbf{Y}_n\}$.

    -   This leads to an overly optimistic estimate and favors complex models.

## Comparing models using KL

-   A better way to estimate elpd is to collect data on a new independent sample that is believed to share the same data generating process as the current sample, and estimate elpd on the new sample.

    -   This is called *out-of-sample validation*.

    -   The problem, of course, is we usually do not have the resources to collect a new sample.

-   Therefore, statisticians have worked hard to find ways to estimate elpd from the current sample, and there are two broad approaches, information criteria and cross-validation.

## Overview of comparison methods

1.  **Information criteria:** AIC, DIC, and WAIC, which estimate the elpd in the current sample, minus a correction factor.

2.  **Cross validation**: A method that splits the current sample into $K$ parts, estimates the parameters in $K − 1$ parts, and estimates the elpd in the remaining 1 part.

-   A special case is when $K = n$ so that each time one uses $n-1$ data points to estimate the model parameters, and estimates the elpd for the observation that was left out. This is called **leave-one-out cross-validation (LOO-CV)**.

## Information criteria

-   Several information criteria have been proposed that do not require fitting the model several times, including AIC, DIC, and WAIC.

-   We will introduce the information criteria, assuming a likelihood $f(\mathbf{Y} | \boldsymbol{\theta})$ for observed data $\mathbf{Y} = (Y_1,\ldots,Y_n)$ with population parameter $\boldsymbol{\theta}$.

-   Information criteria are often presented as **deviance**, defined as, $D(\mathbf{Y}|\boldsymbol{\theta}) = −2 \log f(\mathbf{Y}|\boldsymbol{\theta})$.

-   Ideally, models will have small deviance.

-   However, if a model is too complex it will have small deviance but be unstable (overfitting).

## Akaike information criteria (AIC)

Akaike information criteria (AIC) estimates the elpd as,

$$\widehat{\text{elpd}}_{\text{AIC}} = \log f(\mathbf{Y} | \hat{\boldsymbol{\theta}}_{\text{MLE}}) - p,$$ where $p$ is the number of parameters estimated in the model and $\hat{\boldsymbol{\theta}}_{\text{MLE}}$ is the MLE point estimate.

-   $\text{AIC} = -2\log f(\mathbf{Y} | \hat{\boldsymbol{\theta}}_{\text{MLE}}) + 2p$

-   $p$ is an adjustment for overfitting, but once we go beyond linear models, we cannot simply add $p$.

-   Informative priors tend to reduce the amount of overfitting.

-   Model with smaller AIC are preferred.

## Deviance information criteria (DIC)

Deviance information criteria (DIC) estimates the elpd as,

$$\widehat{\text{elpd}}_{\text{DIC}} = \log f(\mathbf{Y} | \hat{\boldsymbol{\theta}}_{\text{Bayes}}) - p_{\text{DIC}},$$ where $\hat{\boldsymbol{\theta}}_{\text{Bayes}}$ is a Bayesian point estimate, typically a posterior mean, and $p_{\text{DIC}}$ is an estimate of the complexity penalty,

$$p_{\text{DIC}} = 2 \left(\log f(\mathbf{Y} | \hat{\boldsymbol{\theta}}_{\text{Bayes}}) - \mathbb{E}_{\boldsymbol{\theta} | \mathbf{Y}}\left[\log f(\mathbf{Y} | \boldsymbol{\theta}) \right]\right).$$

-   The second term can be estimated as a MC integral.

-   $\text{DIC} = -2\log f(\mathbf{Y} | \hat{\boldsymbol{\theta}}_{\text{Bayes}}) + 2p_{\text{DIC}}.$

## Deviance information criteria (DIC) {.midi}

-   Advantages of DIC:

    -   The effective number of parameters is a useful measure of model complexity.

    -   Intuitively, if there are $p$ parameters and we have uninformative priors then $p_D \approx p$.

    -   However, $p_D \ll p$ if there are strong priors.

-   Disadvantages of DIC:

    -   DIC can only be used to compare models with the same likelihood.

    -   DIC really only applies when the posterior is approximately normal, and will give misleading results when the posterior is far from normality (e.g., bimodal).

## Watanabe-Akaike information criteria (WAIC) {.midi}

Watanabe-Akaike or widely available information criteria (WAIC) estimates the elpd as,

$$\widehat{\text{elpd}}_{\text{WAIC}} = \text{lppd} - p_{\text{WAIC}}.$$

-   The log pointwise predictive density (lppd) is given by, $$\text{lppd} = \log \prod_{i=1}^n f(Y_i | \mathbf{Y}) =  \sum_{i=1}^n \log \int f\left(Y_i | \boldsymbol{\theta}\right)f(\boldsymbol{\theta}| \mathbf{Y}) d\boldsymbol{\theta}.$$

-   lppd can be estimated as, $\sum_{i=1}^n \log \left(\frac{1}{S} \sum_{s = 1}^S f\left(Y_i | \boldsymbol{\theta}^{(s)}\right)\right)$, where $\boldsymbol{\theta}^{(s)}$ are drawn from the posterior.

## WAIC

There are two common estimates of $p_{\text{WAIC}}$, both of which can be estimated using MC samples of the posterior. \begin{align*}
p_{\text{WAIC}_1} &= 2 \sum_{i=1}^n\left(\log \left( \mathbb{E}_{\boldsymbol{\theta} | \mathbf{Y}}\left[f(Y_i | \boldsymbol{\theta})\right]\right) - \mathbb{E}_{\boldsymbol{\theta} | \mathbf{Y}}\left[\log f(Y_i | \boldsymbol{\theta}) \right]\right)\\
p_{\text{WAIC}_2} &= \sum_{i=1}^n \mathbb{V}_{\boldsymbol{\theta} | \mathbf{Y}}\left(\log f\left(Y_i | \mathbf{\theta}\right)\right)
\end{align*}

-   $\text{WAIC} = -2 \text{lppd} + 2p_{\text{WAIC}}.$

## WAIC

-   WAIC has the desirable property of averaging over the posterior distribution, instead of conditioning on a point estimate.

-   $p_{\text{WAIC}}$ can be thought of as an approximation to the number of unconstrained parameters in the model.

-   In practice, $p_{\text{WAIC}_2}$ is often used, since it is theoretically closer to LOO-CV.

## Cross-validation

-   A common approach to compare models is using cross-validation.

-   This is exactly the same procedure used in classical statistics.

-   This operates under the assumption that the *true* model likely produces better out-of-sample predictions than competing models.

-   *Advantages:* Simple, intuitive, and broadly applicable.

-   *Disadvantages:* Slow because it requires several model fits and it is hard to say a difference is statistically significant.

## K-fold cross-validation {.midi}

0.  Split the data into $K$ equally-sized groups.

1.  Set aside group $k$ as test set and fit the model to the remaining $K − 1$ groups.

2.  Make predictions for the test set $k$ based on the model fit to the training data.

3.  Repeat steps 1 and 2 for $k = 1, \dots, K$ giving a predicted value $\widehat{Y}_i$ for all $n$ observations.

4.  Measure prediction accuracy, e.g.,

$$MSE = \frac{1}{n}\sum_{i=1}^n (Y_i - \widehat{Y}_i)^2.$$

## Variants of cross-validation

-   Usually $K$ is either 5 or 10.

-   $K = n$ is called leave-one-out cross-validation (LOO-CV), which is great but slow.

-   The predicted value $\widehat{Y}_i$ can be either the posterior predictive mean or median.

-   Mean squared error (MSE) can be replaced with mean absolute deviation (MAD),

$$MAD = \frac{1}{n}\sum_{i=1}^n |Y_i - \widehat{Y}_i|.$$

## Leave-one-out cross-validation (LOO-CV) {.midi}

-   Assume the data are partitioned into a training set, $\mathbf{Y}_{\text{train}}$ and a holdout set $\mathbf{Y}_{\text{test}}$, thus yielding a posterior distribution $f(\boldsymbol{\theta} | \mathbf{Y}_{\text{train}})$.

-   In the setting of LOO-CV, we have $n$ different $f\left(\boldsymbol{\theta} | \mathbf{Y}_{-i}\right)$.

-   The Bayesian LOO-CV estimate of out-of-sample predictive fit is $$\widehat{\text{elpd}}_{\text{LOO-CV}} = \sum_{i=1}^n \log f\left(Y_i|\mathbf{Y}_{-i}\right),$$ calculated as $$\sum_{i=1}^n \log \left(\frac{1}{S} \sum_{s=1}^S f\left(Y_i|\boldsymbol{\theta}^{(is)}\right)\right),\quad \boldsymbol{\theta}^{(is)} \sim f(\boldsymbol{\theta}|\mathbf{Y}_{-i}).$$

## LOO-CV {.midi}

-   We can write LOO-CV as an information criterion

$$\widehat{\text{elpd}}_{\text{LOO-CV}} = \text{lppd} - p_{LOO-CV},$$

where $$p_{\text{LOO-CV}} = \text{lppd} - \widehat{\text{elpd}}_{\text{LOO-CV}}.$$

-   The estimated number of parameters measures how much in-sample predictive fit overstates out-of-sample performance.

-   This formulation is often called the leave-one-out information criterion (LOO-IC).

## LOO-CV

-   Computing LOO-CV/LOO-IC is computationally demanding.

-   Under some common models there are shortcuts for computing it, however in general these do not exist.

-   WAIC can be treated as a fast approximation of LOO-CV.

-   In Stan, LOO-CV is approximated using the [Pareto smoothed importance sampling (PSIS)](https://arxiv.org/abs/1507.04544) to make the process faster, without having to repeat the process $n$ times.

## Computing WAIC and LOO-CV using Stan

We need to update the **generated quantities** code block.

```{stan output.var = "waic", eval = FALSE}
generated quantities {
  ...
  vector[n] log_lik;
  for (i in 1:n) log_lik[i] = normal_lpdf(Y[i] | X[i, ] * beta, sigma);
}
```

## Let's simulate some data:

```{r}
###True parameters
sigma <- 1.5 # true measurement error
beta <- matrix(c(-1.5, 3, 1), ncol = 1) # true beta

###Simulation settings
n <- 100 # number of observations
n_pred <- 10 # number of predicted observations
p <- length(beta) - 1 # number of covariates

###Simulate data
set.seed(54) # set seed
X <- cbind(1, matrix(rnorm(n * p), ncol = p))
Y <- as.numeric(X %*% beta + rnorm(n, 0, sigma))
X_pred <- cbind(1, matrix(rnorm(n_pred * p), ncol = p))
Y_pred <- as.numeric(X_pred %*% beta + rnorm(n_pred, 0, sigma))
```

## An example model comparison

True Model: $\mathbb{E}[Y_i] = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}$

Model 1: $\mathbb{E}[Y_i] = \beta_0 + \beta_1 x_{i1}$

Model 2: $\mathbb{E}[Y_i] = \beta_0 + \beta_1 x_{i1} + \beta_2 x_{i2}$

## Fit model 1

```{r, eval = FALSE, echo = TRUE}
###Create stan data object
stan_data_model1 <- list(n = n, p = p - 1, Y = Y, X = X[, -3],
                         beta0 = 0, sigma_beta = 10, a = 3,  b = 1,
                         n_pred = n_pred, X_pred = X_pred[, -3])
  
###Compile model separately
stan_model <- stan_model(file = "linear_regression_ppd_log_lik.stan")

###Run model 1 and save
fit_model1 <- sampling(stan_model, data = stan_data_model1, 
                chains = 4, iter = 1000)
saveRDS(fit_model1, file = "linear_regression_ppd_log_lik_fit_model1.rds")
```

```{r, echo=FALSE}
fit_model1 <- readRDS(file = "~/Box Sync/Faculty/Education/biostat725-sp25/course-material/r-objects/linear_regression_ppd_log_lik_fit_model1.rds")
```

## Fit model 2

```{r, eval = FALSE, echo = TRUE}
###Create stan data object
stan_data_model2 <- list(n = n, p = p, Y = Y, X = X,
                         beta0 = 0, sigma_beta = 10, a = 3,  b = 1,
                         n_pred = n_pred, X_pred = X_pred)

###Run model 2 and save
fit_model2 <- sampling(stan_model, data = stan_data_model2, 
                chains = 4, iter = 1000)
saveRDS(fit_model2, file = "linear_regression_ppd_log_lik_fit_model2.rds")
```

```{r, echo=FALSE}
fit_model2 <- readRDS(file = "~/Box Sync/Faculty/Education/biostat725-sp25/course-material/r-objects/linear_regression_ppd_log_lik_fit_model2.rds")
```

## Computing WAIC

-   Begin by extracting the log-likelihood values from the model.

-   We will use the [`loo` package](https://mc-stan.org/loo/).

```{r}
###Load loo package
library(loo)

###Extract log likelihood
log_lik_model1 <- loo::extract_log_lik(fit_model1, parameter_name = "log_lik", merge_chains = TRUE)
log_lik_model2 <- loo::extract_log_lik(fit_model2, parameter_name = "log_lik", merge_chains = TRUE)

###Explore the object
class(log_lik_model1)
dim(log_lik_model1)
```

## Computing WAIC

```{r}
###Compute WAIC for the two models
waic_model1 <- loo::waic(log_lik_model1)
waic_model2 <- loo::waic(log_lik_model2)

###Inspect WAIC for model 1
waic_model1
```

## Computing WAIC

```{r}
###Inspect WAIC for model 2
waic_model2
```

## Computing WAIC

```{r}
###Make a comparison
comp_waic <- loo::loo_compare(list("true" = waic_model2, "misspec" = waic_model1))
print(comp_waic, digits = 2)
print(comp_waic, digits = 2, simplify = FALSE)
```

## Computing LOO-CV/LOO-CI

```{r}
###Compute LOO-IC for the two models
loo_model1 <- loo::loo(log_lik_model1)
loo_model2 <- loo::loo(log_lik_model2)

###Make a comparison
comp <- loo::loo_compare(list("true" = loo_model2, "misspec" = loo_model1))
print(comp, digits = 2)
print(comp, digits = 2, simplify = FALSE)
```

## Prepare for next class

-   Work on [HW 02](https://biostat725-sp26.netlify.app/hw/hw-02), which was just assigned

-   Complete reading to prepare for next Tuesday's lecture

-   Tuesday's lecture: Bayesian Workflow
