---
title: "Multiclass Classification"
author: "Prof. Sam Berchuck"
date: "2025-02-20"
date-format: "MMM DD, YYYY"
footer: "[🔗 BIOSTAT 725 - Spring 2025](https://biostat725-sp25.netlify.app/)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    html-math-method: mathjax
filters:
  - parse-latex
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(knitr)
library(mvtnorm)
library(coda)
library(lars)
library(LaplacesDemon)
library(nimble)
```

## Review of last lecture

-   On Tuesday, we learned about classification using logistic regression.

-   Today, we will focus on multiclass classification: multinomial regression, ordinal regression.

## Multiclass regression

-   Often times one encounters an outcome variable that is nominal and has more than two categories.

-   If there is no inherent rank or order to the variable, we can use multinomial regression. Examples include:

    -   gender (male, female, non-binary),

    -   blood type (A, B, AB, O).

-   If there is an order to the variable, we can use ordinal regression. Examples include:

    -   stages of cancer (stage I, II, III, IV),

    -   pain level (mild, moderate, severe).

## Multinomial random variable

-   Assume an outcome $Y_i \in \{1,\ldots,K\}$ for $i = 1,\ldots,n$.

-   The likelihood in multinomial regression can be written as the following categorical likelihood,

$$f(\mathbf{Y} | \boldsymbol{\beta}) = \prod_{i=1}^n \prod_{j=1}^K P(Y_i = j)^{\delta_{ij}},$$

-   $\delta_{ij} = 1(Y_i = j)$ is the Kronecker delta.

-   Since $Y_i$ is discrete, we only need to specify $P(Y_i = j)$ for all $i$ and $j$.

## Log-linear regression

-   One way to motivate multinomial regression is through a log-linear specification:

$$\log P(Y_i = j) = \mathbf{x}_i\boldsymbol{\beta}_j - \log Z.$$

-   $\boldsymbol{\beta}_j$ is a $j$ specific set of regression parameters.

-   $P(Y_i = j) = \exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}/Z$.

-   $Z$ is a normalizing constant that guarentees that $\sum_{j=1}^K P(Y_i = j) = 1$.

## Finding the normalizing constant

-   We know that,

\begin{align*}
1 &= \sum_{j=1}^K P(Y_i = j) = \frac{1}{Z}\sum_{j=1}^K \exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}\\
&\implies Z = \sum_{j=1}^K \exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}
\end{align*}

## Multinomial probabilities

-   Thus, we have the following,

$$P(Y_i = k) = \frac{\exp\{\mathbf{x}_i\boldsymbol{\beta}_k\}}{\sum_{j = 1}^K\exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}}.$$

-   This function is called the **softmax** function.

-   Unfortunately, this specification is not identifiable.

## Identifiability issue

-   We can add a vector $\mathbf{c}$ to all parameters and get the same result,

\begin{align*}
\frac{\exp\{\mathbf{x}_i(\boldsymbol{\beta}_k + \mathbf{c})\}}{\sum_{j = 1}^K\exp\{\mathbf{x}_i(\boldsymbol{\beta}_j + \mathbf{c})\}} &= \frac{\exp\{\mathbf{x}_i\boldsymbol{\beta}_k\}}{\sum_{j = 1}^K\exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}}.
\end{align*}

-   A common solution is to set: $\boldsymbol{\beta}_K = \mathbf{0}$.

## Updating the probabilities

-   Using the identifiability constraint of $\boldsymbol{\beta}_K = \mathbf{0}$, the probabilities become, \begin{align*}
    P(Y_i = k) &= \frac{\exp\{\mathbf{x}_i\boldsymbol{\beta}_k\}}{1 + \sum_{j = 1}^{K-1}\exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}}, \quad k \in \{1,\ldots,K-1\},\\
    P(Y_i = K) &= \frac{1}{1 + \sum_{j = 1}^{K-1}\exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}}.
    \end{align*}

-   How to interpret the $\boldsymbol{\beta}_k$?

## Deriving the additive log ratio model

-   Using our specification of the probabilities, it can be seen that,

\begin{align*}
P(Y_i = k) &= \frac{\exp\{\mathbf{x}_i\boldsymbol{\beta}_k\}}{1 + \sum_{j = 1}^{K-1}\exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}}\\
&= \left[\frac{1}{1 + \sum_{j = 1}^{K-1}\exp\{\mathbf{x}_i\boldsymbol{\beta}_j\}}\right]\exp\{\mathbf{x}_i\boldsymbol{\beta}_k\}\\
&= P(Y_i = K) \exp\{\mathbf{x}_i \boldsymbol{\beta}_k\}.
\end{align*}

$\implies \log \frac{P(Y_i = k)}{P(Y_i = K)} = \mathbf{x}_i \boldsymbol{\beta}_k, \quad k \in \{1,\ldots,K-1\}$

## Additive log ratio model

-   If outcome $K$ is chosen as reference, the $K − 1$ regression equations are:

$$\log \frac{P(Y_i = k)}{P(Y_i = K)} = \mathbf{x}_i \boldsymbol{\beta}_k, \quad k \in \{1,\ldots, K-1\}$$

-   This formulation is called **additive log ratio**.

-   $\beta_{jk}$ represent the **log-odds** of being in category $k$ relative to the baseline category $K$ with a one-unit change in $X_{ij}$.

    -   $\exp (\beta_{jk})$ is an **odds ratio**.

## Getting back to the likelihood

-   The log-likelihood can be written as,

$$\log f(\mathbf{Y} | \boldsymbol{\beta}) = \sum_{i=1}^n \sum_{j=1}^K \delta_{ij} \log P(Y_i = j).$$

-   The $P(Y_i = j)$ are given by the additive log ratio model.

-   As Bayesians, we only need to specify priors for $\boldsymbol{\beta}_k, k \in \{1,\ldots,K-1\}$.

## Multinomial regression in Stan {.midi}

-   Hard coding the likelihood.

```{stan output.var = "alr", eval = FALSE}
// additive_log_ratio.stan
functions {
  matrix compute_alr_probs(int n, int K, int p, matrix X, matrix beta) {
    matrix[n, K] probs;
    matrix[n, K - 1] expXbeta = exp(X * beta);
    for (i in 1:n) {
      real sum_i = sum(expXbeta[i, ]);
      for (j in 1:K) {
        if (j < K) {
          probs[i, j] = expXbeta[i, j] / (1 + sum_i);
        }
        if (j == K) probs[i, j] = 1 - sum(probs[i, 1:(K - 1)]);
      }
    }
    return probs
  }
}
data {
  int<lower = 1> K;
  int<lower = 1> n;
  int<lower = 1> p;
  array[n] int<lower = 1, upper = K> Y;
  matrix[n, p] X;
  matrix[n, K] delta;
}
parameters {
  matrix[p, K - 1] beta;
}
model {
  matrix[n, K] probs = compute_alr_probs(n, K, p, X, beta);
  for (i in 1:n) {
    for (j in 1:K) {
      target += delta[i, j] * log(probs[i, j]);
    }
  }
  target += normal_lpdf(to_vector(beta) | 0, 10);
}
```

## Multinomial regression in Stan {.midi}

-   Non-identifiable version.

```{stan output.var = "multi_bad", eval = FALSE}
// multi_logit_bad.stan
data {
  int<lower = 1> K;
  int<lower = 1> n;
  int<lower = 1> p;
  array[n] int<lower = 1, upper = K> Y;
  matrix[n, p] X;
}
parameters {
  matrix[p, K] beta;
}
model {
  matrix[n, K] Xbeta = X * beta;
  for (i in 1:n) {
    target += categorical_logit_lpmf(Y[i] | Xbeta[i]')
  }
  target += normal_lpdf(to_vector(beta) | 0, 10);
}
```

[categorical_logit](https://mc-stan.org/docs/functions-reference/bounded_discrete_distributions.html#categorical-logit-glm)

## Multinomial regression in Stan {.midi}

-   Zero identifiability constraint.

```{stan output.var = "multi", eval = FALSE}
// multi_logit.stan
data {
  int<lower = 1> K;
  int<lower = 1> n;
  int<lower = 1> p;
  array[n] int<lower = 1, upper = K> Y;
  matrix[n, p] X;
}
transformed data {
  vector[p] zeros = rep_vector(0, p);
}
parameters {
  matrix[p, K - 1] beta_raw;
}
transformed parameters {
  matrix[p, K] beta = append_col(beta_raw, zeros);
}
model {
  matrix[n, K] Xbeta = X * beta;
  for (i in 1:n) {
    target += categorical_logit_lpmf(Y[i] | Xbeta[i]')
  }
  target += normal_lpdf(to_vector(beta) | 0, 10);
}
generated quantities {
  matrix[p, K - 1] ors = exp(beta_raw);
  vector[n] Y_pred;
  vector[n] log_lik;
  for (i in 1:n) {
    Y_pred[i] = categorical_logit_rng(Xbeta[i]');
    log_lik[i] = categorical_logit_lpmf(Y[i] | Xbeta[i]');
  }
}
```

## Ordinal regression

Let $Y_i \in \{1,\ldots,K\}$ be an ordinal outcome with $K$ categories.

-   The likelihood in ordinal regression is identical to the one from multinomial regression,

$$f(\mathbf{Y} | \boldsymbol{\beta}) = \prod_{i=1}^n \prod_{j=1}^K P(Y_i = j)^{\delta_{ij}}.$$

-   We need to add additional constraints that guarantee ordinality.

## Proportional odds assumption

-   $P(Y_i \leq k)$ is the cumulative probability of $Y_i$ less than or equal to a specific category $k=1,\ldots,K-1$.

-   The odds of being less than or equal to a particular category can be defined as, $$\frac{P(Y\leq k)}{P(Y > k)} \text { for } k=1,\ldots,K-1.$$

-   Not defined for $k = K$, since division by zero is not defined.

## Proportional odds regression

The log odds can then be modeled as follows, $$\log \frac{P(Y_i\leq k)}{P(Y_i > k)} = \text{logit}P(Y_i\leq k) = \alpha_k - \mathbf{x}_i \boldsymbol{\beta}$$

-   Why $-\boldsymbol{\beta}$?

-   $\boldsymbol{\beta}$ is a common regression parameter. For a one-unit increase in $x_{ij}$, $\beta_j$ is the change in log odds of moving to a more severe level of the outcome $Y_i$.

-   $\alpha_k$ for $k = 1,\ldots,K-1$ are $k$-specific intercepts that corresponds to the log odds of moving from level $k$ to $k+1$.

## Understanding the probabilities

-   One can solve for $P(Y_i \leq k), \quad k = 1,\ldots,K-1$),

$$P(Y_i \leq k) = \frac{\exp\{\alpha_k - \mathbf{x}_i\boldsymbol{\beta}\}}{1 + \exp\{\alpha_k - \mathbf{x}_i\boldsymbol{\beta}\}} = \text{expit}(\alpha_k - \mathbf{x}_i\boldsymbol{\beta}).$$

-   $P(Y_i \leq K) = 1$.

The individual probabilities are then given by,

\begin{align*}
P(Y_i = k) &= P(Y_i \leq k) - P(Y_i \leq k-1)\\
&= \text{expit}(\alpha_k - \mathbf{x}_i\boldsymbol{\beta}) - \text{expit}(\alpha_{k-1} - \mathbf{x}_i\boldsymbol{\beta}).
\end{align*}

## A latent variable representation

-   Define a latent variable,

$$Y_i^* = \mathbf{x}_i \boldsymbol{\beta} + \epsilon_i, \quad \epsilon_i \sim \text{Logistic}(0, 1).$$

-   $\mathbb{E}[\epsilon_i] = 0$.

-   $\mathbb{V}(\epsilon_i) = \pi^2/3$.

-   CDF: $P(\epsilon_i \leq x) = \frac{1}{1 + \exp\{-x\}} = \frac{\exp\{x\}}{1 + \exp\{x\}} = \text{expit}(x).$

## Visualizing the latent process

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5
#| layout-ncol: 1
# Parameters for logistic distribution
location <- 0  # Mean of the logistic distribution
scale <- 1  # Scale of the logistic distribution

# Define the thresholds to create ordinal categories
thresholds <- c(-1, 0.5, 1.5)  # These thresholds will create 4 ordinal categories

# Create a sequence of x-values (latent variable values) for plotting
x_vals <- seq(-4, 4, length.out = 1000)

# Calculate the logistic density function (PDF) for the x-values
logistic_density <- dlogis(x_vals, location = location, scale = scale)

# Create a data frame for ggplot
data <- data.frame(x = x_vals, density = logistic_density)

# Create a data frame for shading regions based on thresholds
# Use 'cut' to categorize x_vals into categories based on thresholds
data$category <- cut(data$x, breaks = c(-Inf, thresholds, Inf), 
                     labels = c("Category 1", "Category 2", "Category 3", "Category 4"))


ggplot(data, aes(x = x, y = density)) +
  # Plot the logistic density curve
  geom_line(color = "black", size = 1) +
  # Shade the areas under the curve corresponding to the ordinal categories
  # geom_ribbon(aes(ymin = 0, ymax = density, fill = category), alpha = 0.6) +
  # Customize plot appearance
  scale_fill_manual(values = c("skyblue", "lightgreen", "orange", "salmon")) +
  labs(title = "Logistic(0, 1) Density", 
       x = "Latent Variable (Continuous)", 
       y = "Density") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(-4, 4, by = 1)) +
  theme(legend.position = "top")
```

## Adding thresholds ($c_k$)

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5
#| layout-ncol: 1
# Create the plot for the latent logistic curve with thresholds
ggplot(data, aes(x = x, y = density)) +
  # Plot the logistic density curve without shading
  geom_line(color = "black", size = 1) +
  # Add vertical dashed lines for the thresholds
  geom_vline(xintercept = thresholds, linetype = "dashed", color = "red", size = 1) +
  # Customize plot appearance
  labs(title = "Latent Logistic Curve with Thresholds for K = 4", 
       x = "Latent Variable (Continuous)", 
       y = "Density") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(-4, 4, by = 1)) 
```

## Getting category probabilities

```{r}
#| echo: false
#| fig-align: "center"
#| fig-width: 7
#| fig-height: 5
#| layout-ncol: 1
# Create the plot
ggplot(data, aes(x = x, y = density)) +
  # Plot the logistic density curve
  geom_line(color = "black", size = 1) +
  # Shade the areas under the curve corresponding to the ordinal categories
  geom_ribbon(aes(ymin = 0, ymax = density, fill = category), alpha = 0.6) +
  # Customize plot appearance
  scale_fill_manual(values = c("skyblue", "lightgreen", "orange", "salmon")) +
  labs(title = "Ordinal Regression Visualization with Logistic Density", 
       x = "Latent Variable (Continuous)", 
       y = "Density") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(-4, 4, by = 1)) +
  theme(legend.position = "top")
```

## A latent variable representation

-   Define a set of $K-1$ cut-points, $(c_1,\ldots,c_{K-1}) \in \mathbb{R}^{K-1}$. We also define $c_0 = -\infty, c_K = \infty$.

-   Our ordinal random variable can be generated as,

$$Y_i = \left\{
\begin{matrix*}[l]
1 & c_0 < Y_i^* \leq c_1\\
2 & c_1 < Y_i^* \leq c_2\\
\vdots & \\
K & c_{K-1} < Y_i^* \leq c_K\\
\end{matrix*}
\right.$$

## Equivalence of the two specifications

-   Probabilities under the latent specification: \begin{align*}
    P(Y_i = k) &= P(c_{k-1} < Y_i^* \leq c_k)\\
    &= P(c_{k-1} < \mathbf{x}_i \boldsymbol{\beta} + \epsilon_i \leq c_k)\\
    &= P(\mathbf{x}_i \boldsymbol{\beta} + \epsilon_i \leq c_k) - P(\mathbf{x}_i \boldsymbol{\beta} + \epsilon_i < c_{k-1})\\
    &= P(\epsilon_i \leq c_k - \mathbf{x}_i \boldsymbol{\beta}) - P(\epsilon_i < c_{k-1} - \mathbf{x}_i \boldsymbol{\beta})\\
    &= \text{expit}(c_k - \mathbf{x}_i\boldsymbol{\beta}) - \text{expit}(c_{k-1} - \mathbf{x}_i\boldsymbol{\beta}).
    \end{align*}

-   Equivalency:

    -   $\alpha_k = c_k, \quad k = 1,\ldots, K-1$, assuming that $\alpha_k < \alpha_{k+1}$.

## Ordinal regression using Stan

```{stan output.var = "ordinal", eval = FALSE}
// ordinal.stan
data {
  int<lower = 2> K;
  int<lower = 0> n;
  int<lower = 1> p;
  int<lower = 1, upper = K> Y[n];
  matrix[n, p] X;
}
parameters {
  vector[p] beta;
  ordered[K - 1] alpha;
}
model {
  target += ordered_logistic_glm_lpmf(Y | X, beta, alpha);
  target += normal_lpdf(beta | 0, 10);
  target += normal_lpdf(alpha | 0, 10);
}
generated quantities {
  vector[p] ors = exp(beta);
  vector[n] Y_pred;
  vector[n] log_lik;
  for (i in 1:n) {
    Y_pred[i] = ordered_logistic_rng(X[i, ] * beta, alpha);
    log_lik[i] = ordered_logistic_glm_lpmf(Y[i] | X[i, ], beta, alpha);
  }
}
```

[ordered_logistic_glm_lpmf](https://mc-stan.org/docs/functions-reference/bounded_discrete_distributions.html#ordered-logistic-generalized-linear-model-ordinal-regression)

## Enforcing order in the cutoffs

- In Stan, when you define a parameter as `ordered[K-1] alpha;`, the values of `alpha` are automatically constrained to be strictly increasing. 

- This transformation ensures that the `alpha` values follow the required order, i.e., `alpha[1] < alpha[2] < ... < alpha[K-1]`.

- Stan doesn't sample `alpha` directly but instead works with an **unconstrained** parameter vector, which we will call `gamma`.

## Enforcing order in the cutoffs

```{stan output.var = "order", eval = FALSE}
parameters {
  vector[K - 1] gamma;
}
transformed parameters {
  vector[K - 1] alpha;
  alpha[1] = gamma[1];
  for (j in 2:K) {
    alpha[j] = alpha[j - 1] + exp(gamma[j]);
  }
}
```

- Here, `gamma` represents a vector of independent, unconstrained variables, and the transformation ensures that `alpha` is strictly increasing by construction.

- Luckily we can use [ordered](https://mc-stan.org/docs/reference-manual/transforms.html#ordered-vector), since Stan takes care of this (including the Jacobian) in the background.

## Prepare for next class

-   Work on [HW 03](https://biostat725-sp25.netlify.app/hw/hw-03).

-   Complete reading to prepare for next Tuesday's lecture

-   Tuesday's lecture: Missing data
