---
title: "Longitudinal Data"
author: "Prof. Sam Berchuck"
date: "2026-02-26"
date-format: "MMM DD, YYYY"
footer: "[🔗 BIOSTAT 725 - Spring 2026](https://biostat725-sp26.netlify.app/)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    html-math-method: mathjax
filters:
  - parse-latex
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r, echo=FALSE}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(knitr)
library(mvtnorm)
library(coda)
library(lars)
library(LaplacesDemon)
library(nimble)
library(mice)
library(bayesplot)
```

## Review of last lecture

-   During our last lecture, we introduced correlated (or dependent) data sources.

-   We discussed the idea of accounting for dependencies within a group using group-specific parameters.

-   We introduced the random intercept model and studied the induced correlation (forced to be positive) in the marginal model.

-   Today we will look at longitudinal data and introduce a simple model that accounts for group-level changes.

## Longitudinal Data

Repeated measurements taken over time from the same subjects. Examples include:

-   **Monitor Disease Progression**: Track how diseases evolve, such as diabetes or glaucoma.

-   **Evaluate Treatments**: Understand how interventions work over time.

-   **Personalized Health Insights**: Capture individual health trajectories for personalized care.

-   **Study Long-Term Effects**: Evaluate the long-term outcomes of medical treatments or behaviors.

## Example: Glaucoma Disease Progression

Imagine we are tracking mean deviation (MD, dB), a key measure of visual field loss in glaucoma patients, over time.

-   Multiple measurements of MD for each patient across several years.

-   We're interested in glaucoma progression, which is defined as the rate of change in MD over time (dB/year).

-   Define $Y_{it}$ as the MD value for eye $i$ ($i = 1,\ldots,n$) at time $t$ ($t = 1,\ldots,n_i$) and the time of each observation as $X_{it}$ with $X_{i1} = 0$.

## Rotterdam data

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
glaucoma_longitudinal <- readRDS(file = "/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/data/glaucoma/glaucoma_longitudinal.rds")
n_eyes <- length(unique(glaucoma_longitudinal$eye_id))
set.seed(54)
ggplot(glaucoma_longitudinal[glaucoma_longitudinal$eye_id %in% sample(1:n_eyes, 12), ], aes(x = time, y = mean_deviation)) +
  geom_point() +
  geom_line() +
  facet_wrap(vars(eye_id), scales = "free") +
  labs(y = "Mean deviation (dB)",
       x = "Time from baseline visit (years)",
       subtitle = "Longitudinal series of visual fields for 12 example patients")
```

## Treating Eyes Separately

We can model each eye **separately** using OLS (this is a form of longitudinal analysis!). For $t = 1,\ldots,n_i$, the model is:

$$Y_{it} = \beta_{0i} + X_{it} \beta_{1i} + \epsilon_{it}, \quad \epsilon_{it} \stackrel{iid}{\sim} N(0, \sigma_i^2).$$

Where:

-   $\beta_{0i}$ is the intercept for eye $i$.

-   $\beta_{1i}$ is the slope for eye $i$ (i.e., disease progression).

-   $\sigma_i^2$ is the residual error for eye $i$.

## OLS regression

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 5
#| fig-align: "center"
glaucoma_longitudinal <- readRDS(file = "/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/data/glaucoma/glaucoma_longitudinal.rds")
n_eyes <- length(unique(glaucoma_longitudinal$eye_id))
set.seed(54)
ggplot(glaucoma_longitudinal[glaucoma_longitudinal$eye_id %in% sample(1:n_eyes, 12), ], aes(x = time, y = mean_deviation)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(vars(eye_id), scales = "free") +
  labs(y = "Mean deviation (dB)",
       x = "Time from baseline visit (years)",
       subtitle = "Longitudinal series of visual fields for 12 example patients")
```

## Treating Eyes Separately

-   Fitting OLS separately allows **each eye** to have a unique intercept and slope, which of course is consistent with the data generating process.

-   However, this can lead to eye-specific intercepts and slopes that are not realistic (consider OLS regression with very few data points).

-   Estimating eye-specific intercepts and slopes within the context of the whole study sample should shrink extreme values toward the population average.

## Subject-specific intercepts and slopes

For $i = 1,\ldots,n$ and $t=1,\ldots,n_i$, we can write the model:

\begin{align*}
Y_{it} &= \beta_{0i} + X_{it} \beta_{1i} + \epsilon_{it}, \quad \epsilon_{it} \stackrel{iid}{\sim} N(0, \sigma^2),\\
\beta_{0i} &= \beta_0 + \theta_{0i},\\
\beta_{1i} &= \beta_1 + \theta_{1i}.
\end{align*}

**Population Parameters**:

-   $\beta_0$ is the population intercept (i.e., average MD value in the population at time zero).

-   $\beta_1$ is the population slope (i.e., average disease progression).

-   $\sigma^2$ is the population residual error.

## Subject-specific intercepts and slopes

For $i = 1,\ldots,n$ and $t=1,\ldots,n_i$, we can write the model:

\begin{align*}
Y_{it} &= \beta_{0i} + X_{it} \beta_{1i} + \epsilon_{it}, \quad \epsilon_{it} \stackrel{iid}{\sim} N(0, \sigma^2),\\
\beta_{0i} &= \beta_0 + \theta_{0i},\\
\beta_{1i} &= \beta_1 + \theta_{1i}.
\end{align*}

**Subject-Specific Parameters**:

-   $\theta_{0i}$ is the subject-specific deviation from the intercept for eye $i$.

-   $\theta_{1i}$ is the subject-specific deviation from the slope for eye $i$.

## Subject-specific intercepts and slopes

For $i = 1,\ldots,n$ and $t=1,\ldots,n_i$, we can write the model:

\begin{align*}
Y_{it} &= \beta_{0i} + X_{it} \beta_{1i} + \epsilon_{it}, \quad \epsilon_{it} \stackrel{iid}{\sim} N(0, \sigma^2),\\
\beta_{0i} &= \beta_0 + \theta_{0i},\\
\beta_{1i} &= \beta_1 + \theta_{1i}.
\end{align*}

**Key Advantage**:

-   This model defines subject-specific estimates of $\beta_{0i}$ and $\beta_{1i}$ relative to the population average, preventing overfitting and making the estimates more stable.

-   **Shrinks** subject-specific parameters to the population average.

## Linear Mixed Model

The subject-specific intercepts and slope model can be seen as a special case of the linear mixed model (LMM). For $i = 1,\ldots,n$, LMM is defined as:

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

-   $\mathbf{Y}_i = (Y_{i1},\ldots,Y_{in_i})$ are subject-level observations.

-   $Y_{it}$ is the $t$th observation in subject $i$.

-   $\boldsymbol{\epsilon}_i = (\epsilon_{i1},\ldots,\epsilon_{in_i})$, such that $\epsilon_{it} \stackrel{iid}{\sim} N(0,\sigma^2)$.

## Linear Mixed Model

For $i = 1,\ldots,n$, the linear mixed model (LMM) is given by:

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

-   $\mathbf{X}_i$ is an $(n_i \times p)$-dimensional matrix with row $\mathbf{x}_{it}$ (intercept is incorporated).

-   $\mathbf{x}_{it}$ contains variables that are assumed to relate to the outcome only at a population-level.

-   $p$ is the number of population-level variables.

## Linear Mixed Model

For $i = 1,\ldots,n$, the linear mixed model (LMM) is given by:

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

-   $\mathbf{Z}_i$ is an $(n_i \times q)$-dimensional matrix with row $\mathbf{z}_{it}$ (intercept is incorporated).

-   $\mathbf{z}_{it}$ contains variables that are assumed to relate to the outcome with varying effects at a subject-level.

-   $q$ is the number of subject-level variables.

## Linear Mixed Model

For $i = 1,\ldots,n$, the linear mixed model (LMM) is given by:

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

-   $\boldsymbol{\beta}$ is a $p$-dimensional vector of population-level parameters (or fixed effects).

-   $\boldsymbol{\theta}_i$ is a $q$-dimensional vector of group-level parameters (or random effects).

-   $\sigma^2$ is a population-level parameter that measures residual error.

## Recover the Random Intercept Model

For $i = 1,\ldots,n$, the linear mixed model (LMM) is given by:

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

Suppose that $\mathbf{z}_{it} = 1 \forall i,t$. Then we get

\begin{align*}
Y_{it} &= \mathbf{x}_{it} \boldsymbol{\beta} + \mathbf{z}_{it}\boldsymbol{\theta}_{i} + \epsilon_{it}, \quad \epsilon_{it} \stackrel{iid}{\sim} N(0,\sigma^2)\\
&= \mathbf{x}_{it} \boldsymbol{\beta} + \theta_{i} + \epsilon_{it}.
\end{align*}

-   LMM is a general form of the random intercept model.

## Random Slope and Intercept Model

For $i = 1,\ldots,n$, the linear mixed model (LMM) is given by:

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

Suppose that $\mathbf{x}_{it} = \mathbf{z}_{it} = (1, X_{it})$, such that $p = q = 2$. Then,

\begin{align*}
Y_{it} &= \mathbf{x}_{it} \boldsymbol{\beta} + \mathbf{z}_{it}\boldsymbol{\theta}_{i} + \epsilon_{it}, \quad \epsilon_{it} \stackrel{iid}{\sim} N(0,\sigma^2)\\
&= \beta_0 + \beta_1 X_{it} + \theta_{0i} + \theta_{1i} X_{it} + \epsilon_{it}\\
&= (\beta_0 + \theta_{0i}) + (\beta_1 + \theta_{1i}) X_{it} + \epsilon_{it}.
\end{align*}

where $\boldsymbol{\beta} = (\beta_0,\beta_1)$ and $\boldsymbol{\theta}_i = (\theta_{0i},\theta_{1i})$.

## Prior Specification {.midi}

One choice could be to specify independent priors for the subject-specific intercepts and slopes:

\begin{align*}
\theta_{0i} &\stackrel{iid}{\sim} N(0, \tau_0^2)\\
\theta_{1i} &\stackrel{iid}{\sim} N(0, \tau_1^2).
\end{align*}

-   This is the same assumption we made last lecture, where we assume a normal distribution centered at zero with some variance that reflects variability across subjects.

-   Often times this assumption is oversimplified. For example in glaucoma progression, we often assume that if someone has a higher baseline MD they will a more negative slope (i.e., negative correlation).

## Prior Specification {.midi}

We can instead model the subject-specific parameters as correlated themselves using a bi-variate normal distribution. Define $\boldsymbol{\theta}_i = (\theta_{0i},\theta_{1i})^\top$ and then $\boldsymbol{\theta}_i \stackrel{iid}{\sim} N_2(\mathbf{0}_2,\boldsymbol{\Sigma})$.

$$\boldsymbol{\Sigma} = \begin{bmatrix}
    \tau_{0}^2 & \tau_{01}\\
    \tau_{01} & \tau_1^2\\
  \end{bmatrix}.$$

-   $\tau_{01} = \rho \tau_0 \tau_1$.

-   $\rho$ is the correlation between the subject-specific intercepts and slopes.

**Let's talk about efficient ways to generate multivariate random variables!**

## Generating Multivariate Normal RNGs {.midi}

Suppose we would like to generate samples of a random variable $\mathbf{x}_i \stackrel{iid}{\sim} N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})$.

To sample efficiently, we can decompose the covariance structure:

\begin{align*}
\boldsymbol{\Sigma} &= \begin{bmatrix}
    \tau_{0}^2 & \rho \tau_0 \tau_1\\
    \rho \tau_0 \tau_1 & \tau_1^2\\
  \end{bmatrix}\\
&= \begin{bmatrix}
    \tau_{0} & 0\\
    0 & \tau_1\\
  \end{bmatrix}  \begin{bmatrix}
    1 & \rho\\
    \rho & 1\\
  \end{bmatrix}  \begin{bmatrix}
    \tau_{0} & 0\\
    0 & \tau_1\\
  \end{bmatrix}\\
&=  \mathbf{D} \boldsymbol{\Phi} \mathbf{D}.
\end{align*}

-   $\mathbf{D}$ is a $p$-dimensional matrix with the standard deviations on the diagonal.

-   $\boldsymbol{\Phi}$ is the correlation matrix.

## Generating Multivariate Normal RNGs

We can further decompose the covariance by computing the cholesky decomposition of the correlation matrix:

\begin{align*}
\boldsymbol{\Sigma} &=  \mathbf{D} \boldsymbol{\Phi} \mathbf{D}\\
&= \mathbf{D} \mathbf{L} \mathbf{L}^\top \mathbf{D},
\end{align*} where $\mathbf{L}$ is the lower triangular Cholesky decomposition for $\boldsymbol{\Phi}$, such that $\boldsymbol{\Phi} = \mathbf{L} \mathbf{L}^\top$.

## Generating Multivariate Normal RNGs

We can generate samples $\mathbf{x}_i \stackrel{iid}{\sim} N_2(\boldsymbol{\mu}, \boldsymbol{\Sigma})$ using the following approach:

$$\mathbf{x}_i = \boldsymbol{\mu} + \mathbf{D} \mathbf{L} \mathbf{z}_i,$$

where $\mathbf{z}_i = (z_{0i},z_{1i})$ and $z_{ij} \stackrel{iid}{\sim} N(0,1)$, so that $\mathbb{E}[\mathbf{z}_i] = \mathbf{0}_2$ and $\mathbb{C}(\mathbf{z}_i) = \mathbf{I}_2$.

\begin{align*}
\mathbb{E}[\boldsymbol{\mu} + \mathbf{D}\mathbf{L}\mathbf{z}_i] &= \boldsymbol{\mu} +  \mathbf{D}\mathbf{L}\mathbb{E}[\mathbf{z}_i] = \boldsymbol{\mu}\\
\mathbb{C}(\boldsymbol{\mu} + \mathbf{D}\mathbf{L}\mathbf{z}_i) &= \mathbf{D}\mathbf{L}\mathbb{C}(\mathbf{z}_i)\left(\mathbf{D}\mathbf{L}\right)^\top \\
&= \mathbf{D}\mathbf{L}\mathbf{L}^\top\mathbf{D}\\
&=\boldsymbol{\Sigma}.
\end{align*}

## Generating Multivariate Normal RNGs {.midi}

```{r}
Sigma <- matrix(c(3, 1, 1, 3), nrow = 2, ncol = 2, byrow = TRUE)
mu <- matrix(c(2, 5), ncol = 1)
D <- matrix(0, nrow = 2, ncol = 2)
diag(D) <- diag(sqrt(Sigma))
Phi <- cov2cor(Sigma)
L <- t(chol(Phi))
n_samples <- 1000
z <- matrix(rnorm(n_samples * 2), nrow = 2, ncol = n_samples)
mu_mat <- matrix(rep(mu, n_samples), nrow = 2, ncol = n_samples) 
X <- mu_mat + D %*% L %*% z
apply(X, 1, mean)
cov(t(X))
```

## Conditional specification

For the conditional specification, we can write the model at the observational level, $Y_{it}$. This is because conditionally on the $\boldsymbol{\theta}_i$, $Y_{it}$ and $Y_{it'}$ are independent.

For $i$ ($i = 1,\ldots,n$) and $t$ ($t = 1,\ldots, n_i$), the model is:

\begin{align*}
Y_{it} | \boldsymbol{\Omega}, \boldsymbol{\theta}_i &\stackrel{iid}{\sim} N\left((\beta_{0} + \theta_{0i}) + (\beta_1 + \theta_{0i}) X_{it}, \sigma^2\right),\\
\boldsymbol{\theta}_i | \boldsymbol{\Sigma} &\stackrel{iid}{\sim} N_2(\mathbf{0}_2,\boldsymbol{\Sigma})\\
\boldsymbol{\Omega} &\sim f(\boldsymbol{\Omega}),
\end{align*}

where $\boldsymbol{\Omega} = (\beta_0, \beta_1, \sigma^2, \boldsymbol{\Sigma})$.

## Conditional Specification

-   Moments for the Conditional Model:

\begin{align*}
\mathbb{E}[Y_{it} | \boldsymbol{\Omega},\boldsymbol{\theta}_i] &= (\beta_0 + \theta_{0i}) + (\beta_1 + \theta_{1i}) X_{it}\\
\mathbb{V}(Y_{it} | \boldsymbol{\Omega},\boldsymbol{\theta}_i) &= \sigma^2\\
\mathbb{C}(Y_{it}, Y_{jt'} | \boldsymbol{\Omega},\boldsymbol{\theta}_i,\boldsymbol{\theta}_{t'}) &= 0,\quad \forall i,j,t,t'
\end{align*}

<!-- ## Understanding the conditional mean -->

<!-- \begin{align*} -->

<!-- \mathbb{E}[Y_{it} | \boldsymbol{\Omega},\boldsymbol{\theta}_i] &= \beta_{0i} + \beta_{1i} X_{it}\\ -->

<!-- &= (\beta_0 + \theta_{0i}) + (\beta_1 + \theta_{1i}) X_{it}\\ -->

<!-- &= \beta_0 + \beta_1 X_{it} + \theta_{0i} + \theta_{1i} X_{it}\\ -->

<!-- &= \mathbf{x}_{it} \boldsymbol{\beta} + \mathbf{x}_{it} \boldsymbol{\theta}_i, -->

<!-- \end{align*} -->

<!-- where $\mathbf{x}_{it} = (1, X_{it})$ and $\boldsymbol{\beta} = (\beta_0, \beta_1)$. -->

## Conditional Specification

-   Define $\mathbf{Y}_i = (Y_{i1},\ldots,Y_{in_i})$ and $\mathbf{Y} = (\mathbf{Y}_1,\ldots,\mathbf{Y}_n)$.

-   The posterior for the conditional model can be written as:

\begin{align*}
f(\boldsymbol{\Omega}, \boldsymbol{\theta} | \mathbf{Y}) &\propto f(\mathbf{Y}, \boldsymbol{\Omega}, \boldsymbol{\theta})\\
&= f(\mathbf{Y} | \boldsymbol{\Omega}, \boldsymbol{\theta}) f(\boldsymbol{\theta} | \boldsymbol{\Omega})f(\boldsymbol{\Omega})\\
&=  \prod_{i=1}^n \prod_{t = 1}^{n_i} f(Y_{it} | \boldsymbol{\Omega}, \boldsymbol{\theta})  \prod_{i=1}^n f(\boldsymbol{\theta}_i | \boldsymbol{\Sigma}) f(\boldsymbol{\Omega}),
\end{align*} where $\boldsymbol{\theta} = (\boldsymbol{\theta}_1,\ldots,\boldsymbol{\theta}_n)$.

<!-- ## Marginal Specification {.midi} -->

<!-- To derive a marginal model it is useful to write the model at the level of the independent observations, $\mathbf{Y}_i$. -->

<!-- $$\mathbf{Y}_i = \begin{bmatrix} -->

<!--     Y_{i1}\\ -->

<!--     Y_{i2}\\ -->

<!--     \vdots\\ -->

<!--     Y_{in_i} -->

<!--   \end{bmatrix} =  -->

<!--   \begin{bmatrix} -->

<!--     \mathbf{x}_{i1} \boldsymbol{\beta} + \mathbf{x}_{i1} \boldsymbol{\theta}_i + \epsilon_{i1}\\ -->

<!--     \mathbf{x}_{i2} \boldsymbol{\beta} + \mathbf{x}_{i2} \boldsymbol{\theta}_i + \epsilon_{i2}\\ -->

<!--     \vdots \\ -->

<!--     \mathbf{x}_{in_i} \boldsymbol{\beta} + \mathbf{x}_{in_i} \boldsymbol{\theta}_i + \epsilon_{in_i} -->

<!--   \end{bmatrix} = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{X}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i,$$ where $\mathbf{x}_{it} = (1, X_{it})$, $\mathbf{X}_i$ is an $n_i \times 2$ dimensional matrix with rows $\mathbf{x}_{it}$. -->

<!-- -   $\boldsymbol{\epsilon}_i = (\epsilon_{i1},\ldots,\epsilon_{in_i}) \stackrel{ind}{\sim} N(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i})$. -->

## Marginal Specification

The LMM model is given by:

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}_i + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

-   Moments for the Marginal Model:

\begin{align*}
\mathbb{E}[\mathbf{Y}_{i} | \boldsymbol{\Omega}] &= \mathbf{X}_i\boldsymbol{\beta}\\
\mathbb{V}(\mathbf{Y}_{i} | \boldsymbol{\Omega}) &= \mathbf{Z}_i \boldsymbol{\Sigma} \mathbf{Z}_i^\top + \sigma^2 \mathbf{I}_{n_i} = \boldsymbol{\Upsilon}_i\\
\mathbb{C}(\mathbf{Y}_{i}, \mathbf{Y}_{i'} | \boldsymbol{\Omega}) &= \mathbf{0}_{n_i \times n_i},\quad i \neq i'.
\end{align*}

## Marginal Specification

For $i = 1,\ldots,n$, \begin{align*}
\mathbf{Y}_{i} | \boldsymbol{\Omega} &\stackrel{ind}{\sim} N(\mathbf{X}_i\boldsymbol{\beta},\boldsymbol{\Upsilon}_i)\\
\boldsymbol{\Omega} &\sim f(\boldsymbol{\Omega}),
\end{align*} where $\boldsymbol{\Omega} = (\boldsymbol{\beta},\sigma,\boldsymbol{\Sigma})$ are the population parameters.

## Recovering the Subject-Specific Parameters {.midi}

-   We can still recover the $\boldsymbol{\theta}_i$ when we fit the marginal model, we only need to compute $f(\boldsymbol{\theta}_i | \mathbf{Y}_i,\boldsymbol{\Omega})$ for all $i$.

-   We can obtain this full conditional by specifying the joint distribution,

$$f\left(\begin{bmatrix}
    \mathbf{Y}_i\\
    \boldsymbol{\theta}_i
  \end{bmatrix} \Bigg| \boldsymbol{\Omega}\right) = N\left(\begin{bmatrix}
    \mathbf{X}_i \boldsymbol{\beta} \\
    \mathbf{0}_{n_1}
  \end{bmatrix}, \begin{bmatrix}
    \boldsymbol{\Upsilon}_i & \mathbf{Z}_i\boldsymbol{\Sigma}\\
    \boldsymbol{\Sigma} \mathbf{Z}_i^\top & \boldsymbol{\Sigma}
  \end{bmatrix}\right).$$

We can then use the [conditional specification of a multivariate normal](https://en.wikipedia.org/wiki/Multivariate_normal_distribution#Conditional_distributions) to find, $f(\boldsymbol{\theta}_i | \mathbf{Y}_i, \boldsymbol{\Omega}) = N(\mathbb{E}_{\boldsymbol{\theta}_i},\mathbb{V}_{\boldsymbol{\theta}_i})$, where

\begin{align*}
\mathbb{E}_{\boldsymbol{\theta}_i} &= \mathbf{0}_{n_i} + \boldsymbol{\Sigma} \mathbf{Z}_i^\top \boldsymbol{\Upsilon}_i^{-1} (\mathbf{Y}_i - \mathbf{X}_i \boldsymbol{\beta})\\
\mathbb{V}_{\boldsymbol{\theta}_i} &= \boldsymbol{\Sigma} - \boldsymbol{\Sigma} \mathbf{Z}_i^\top \boldsymbol{\Upsilon}_i^{-1} \mathbf{Z}_i\boldsymbol{\Sigma}.
\end{align*}

## Marginal Specification

It is not as straightforward to gain intuition behind the induced correlation structure, but we can shed some light by studying the scalar form of the covariance:

\begin{align*}
\mathbb{V}(Y_{it}| \boldsymbol{\Omega}) &= \tau_0^2 + 2 \tau_{01} X_{it}^2 + \tau_1^2 X_{it}^2 + \sigma^2,\\
\mathbb{C}(Y_{it}, Y_{it'} | \boldsymbol{\Omega}) &= \tau_0^2 - \rho \tau_0 \tau_1 (X_{it} - X_{it'}) + \tau_1^2 X_{it} X_{it'}.
\end{align*}

## Visualizing the dependency

$\tau_0 = 1,\tau_1 = 1,\sigma^2 =2, \rho = 0.5$

```{r}
#| fig-align: "center"
#| fig-width: 9
#| fig-height: 8
#| echo: false
tau0 <- 1
tau1 <- 1
sigma2 <- 2
rho <- 0.5
tau01 <- rho * tau0 * tau1
x <- seq(0, 1, length.out = 25)
x_s <- expand.grid(x, x)
correlation <- numeric(length = nrow(x_s))
for (i in 1:nrow(x_s)) {
  xs <- x_s[i, ]
  covariance <- tau0^2 + tau01 * xs[1] + tau01 * xs[2] + tau1^2 * prod(xs)
  var1 <- tau0^2 + 2 * tau01 * xs[1]^2 + tau1^2 * xs[1]^2 + sigma2
  var2 <- tau0^2 + 2 * tau01 * xs[2]^2 + tau1^2 * xs[2]^2 + sigma2
  correlation[i] <- as.numeric(covariance / sqrt(var1 * var2))
}
dat_fig <- data.frame(
  x_s,
  correlation
)
ggplot(dat_fig, aes(x = Var1, y = Var2, fill = correlation)) + 
  geom_tile() +  # Use geom_tile for heatmap-like visualization
  scale_fill_viridis_c() +  # Optional color scale
  coord_fixed() +  # Fix aspect ratio
  labs(title = "Correlation Heatmap",
       x = expression(x[it]),
       y = expression(x[is]),
       fill = "Correlation") + # Optional title 
  theme_minimal()
```

## Visualizing the dependency

$\tau_0 = 1,\tau_1 = 1,\sigma^2 =2, \rho = -0.5$

```{r}
#| fig-align: "center"
#| fig-width: 9
#| fig-height: 8
#| echo: false
tau0 <- 1
tau1 <- 1
sigma2 <- 2
rho <- -0.5
tau01 <- rho * tau0 * tau1
x <- seq(0, 1, length.out = 25)
x_s <- expand.grid(x, x)
correlation <- numeric(length = nrow(x_s))
for (i in 1:nrow(x_s)) {
  xs <- x_s[i, ]
  covariance <- tau0^2 + tau01 * xs[1] + tau01 * xs[2] + tau1^2 * prod(xs)
  var1 <- tau0^2 + 2 * tau01 * xs[1]^2 + tau1^2 * xs[1]^2 + sigma2
  var2 <- tau0^2 + 2 * tau01 * xs[2]^2 + tau1^2 * xs[2]^2 + sigma2
  correlation[i] <- as.numeric(covariance / sqrt(var1 * var2))
}
dat_fig <- data.frame(
  x_s,
  correlation
)
ggplot(dat_fig, aes(x = Var1, y = Var2, fill = correlation)) + 
  geom_tile() +  # Use geom_tile for heatmap-like visualization
  scale_fill_viridis_c() +  # Optional color scale
  coord_fixed() +  # Fix aspect ratio
  labs(title = "Correlation Heatmap",
       x = expression(x[it]),
       y = expression(x[is]),
       fill = "Correlation") + # Optional title 
  theme_minimal()
```

## Marginal Specification

-   The posterior for the conditional model can be written as:

\begin{align*}
f(\boldsymbol{\Omega} | \mathbf{Y}) &\propto f(\mathbf{Y}, \boldsymbol{\Omega})\\
&= f(\mathbf{Y} | \boldsymbol{\Omega})f(\boldsymbol{\Omega})\\
&=  \prod_{i=1}^n f(\mathbf{Y}_{i} | \boldsymbol{\Omega}) f(\boldsymbol{\Omega}).
\end{align*}

## Specifying a Prior Distribution for $\boldsymbol{\Omega}$

-   We must set a prior for $f(\boldsymbol{\Omega}) = f(\boldsymbol{\beta}) f(\sigma) f(\boldsymbol{\Sigma})$.

-   We can place standard priors on $\boldsymbol{\beta}$ and $\sigma$.

-   $\boldsymbol{\Sigma}$ is a covariance (i.e., positive definite matrix), so we must be careful here.

-   It is natural to place a prior on the decomposition, $\boldsymbol{\Sigma} = \mathbf{D} \mathbf{L} \mathbf{L}^\top \mathbf{D}$.

    -   For each of the standard deviations $(\tau_0,\tau_1)$ we can place standard priors for scales (e.g., half-normal).

    -   For $\mathbf{L}$ we can place a [Lewandowski-Kurowicka-Joe (LKJ)](https://mc-stan.org/docs/functions-reference/correlation_matrix_distributions.html#cholesky-lkj-correlation-distribution) distribution, $\mathbf{L} \sim LKJ(\eta)$.

## What Does the LKJ Prior Do?

-   The LKJ prior allows you to model the correlation structure in a flexible and non-informative way.

-   It is defined by a single parameter, $\eta > 0$, which controls the concentration of the prior.

    -   When $(\eta = 1)$, it is an uninformative prior (i.e., uniform on the correlations).

    -   When $(\eta > 1)$, the prior favors more highly correlated random effects.

    -   When $(\eta < 1)$, the prior favors weaker correlations.

-   When $q=2$, $\eta = 1$ is equivalent to $\rho \sim \text{Uniform}(-1,1)$.

## LKJ Prior Formula

-   The LKJ prior on a correlation matrix $\mathbf{L}$ is given by:

$$f(\mathbf{L} | \eta) \propto \prod_{j = 2}^q L_{jj}^{q-j+2\eta-2}.$$

Where:

-   $\eta$ is the concentration parameter.

-   $q$ is the size of the correlation matrix.

-   $L_{jk}$ is the observation in the $j$th row and $k$th column of $\mathbf{L}$.

## LKJ Prior in Stan

```{stan output.var = "chol", eval = FALSE}
parameters {
  cholesky_factor_corr[2] L;  // Cholesky factor of correlation matrix
}
model {
  L ~ lkj_corr_cholesky(eta);
}
```

## Stan code for independent intercept and slope

```{stan output.var = "ind", eval = FALSE}
// lmm-independent.stan
data {
  int<lower = 1> n;
  int<lower = 1> N;
  vector[N] Time;
  vector[N] MD;
  int<lower = 1, upper = n> Ids[N];
}
parameters {
  real beta0;
  real beta1;
  real<lower = 0> sigma;
  vector[n] z0;
  vector[n] z1;
  real<lower = 0> tau0;
  real<lower = 0> tau1;
}
transformed parameters {
  vector[n] theta0;
  vector[n] theta1;
  theta0 = tau0 * z0;
  theta1 = tau1 * z1;
}
model {
  // likelihood
  vector[N] mu;
  for (i in 1:N) {
    mu[i] = (beta0 + theta0[Ids[i]]) + (beta1 + theta1[Ids[i]]) * Time[i];
  }
  target += normal_lpdf(MD | mu, sigma);
  // subject-specific parameters
  target += std_normal_lpdf(z0);
  target += std_normal_lpdf(z1);
  // population parameters
  target += normal_lpdf(beta0 | 0, 3);
  target += normal_lpdf(beta1 | 0, 3);
  target += normal_lpdf(sigma | 0, 3);
  target += normal_lpdf(tau0 | 0, 3);
  target += normal_lpdf(tau1 | 0, 3);
}
```

## Stan code for conditional LMM

```{stan output.var = "cond", eval = FALSE}
// lmm-conditional.stan
data {
  int<lower = 1> N;
  int<lower = 1> n;
  int<lower = 1> p;
  int<lower = 1> q;
  matrix[N, p] X;
  matrix[N, q] Z;
  vector[N] Y;
  int<lower = 1, upper = n> Ids[N];
  real<lower = 0> eta;
}
parameters {
  vector[p] beta;
  real<lower = 0> sigma;
  matrix[q, n] z;
  vector<lower = 0>[q] tau;
  cholesky_factor_corr[q] L;
}
transformed parameters {
  matrix[q, n] theta;
  theta = diag_pre_multiply(tau, L) * z;
}
model {
  // likelihood
  vector[N] mu;
  for (i in 1:N) {
    mu[i] = X[i, ] * beta + Z[i, ] * theta[, Ids[i]];
  }
  target += normal_lpdf(Y | mu, sigma);
  // subject-specific parameter
  target += std_normal_lpdf(to_vector(z));
  // population parameters
  target += normal_lpdf(beta | 0, 3);
  target += normal_lpdf(sigma | 0, 3);
  target += normal_lpdf(tau | 0, 3);
  target += lkj_corr_cholesky_lpdf(L | eta);
}
generated quantities {
  corr_matrix[q] Phi = L * transpose(L);
  real rho = Phi[1, 2];
  vector[n] subject_intercepts = beta[1] + to_vector(theta[1, ]);
  vector[n] subject_slopes = beta[2] + to_vector(theta[2, ]);
  vector[N] Y_pred;
  vector[N] log_lik;
  vector[N] mu;
  for (i in 1:N) {
    mu[i] = X[i, ] * beta + Z[i, ] * theta[, Ids[i]];
    Y_pred[i] = normal_rng(mu[i], sigma);
    log_lik[i] = normal_lpdf(Y[i] | mu[i], sigma);
  }
}
```

## Stan code for marginal LMM

Need [ragged data structure](https://mc-stan.org/docs/stan-users-guide/sparse-ragged.html#ragged-data-structs.section).

```{stan output.var = "cond", eval = FALSE}
// lmm-marginal.stan
data {
  int<lower = 1> N;
  int<lower = 1> n;
  int<lower = 1> p;
  int<lower = 1> q;
  matrix[N, p] X;
  matrix[N, q] Z;
  vector[N] Y;
  int<lower = 1> n_is[n];
  real<lower = 0> eta;
}
parameters {
  vector[p] beta;
  real<lower = 0> sigma;
  vector<lower = 0>[q] tau;
  cholesky_factor_corr[q] L;
}
transformed parameters {
  cov_matrix[q] Sigma;
  Sigma = diag_pre_multiply(tau, L) * transpose(diag_pre_multiply(tau, L));
}
model {
  // evaluate the likelihood for the marginal model using ragged data structure
  int pos;
  pos = 1;
  for (i in 1:n) {
    int n_i = n_is[i];
    vector[n_i] Y_i = segment(Y, pos, n_i);
    matrix[n_i, p] X_i;
    matrix[n_i, q] Z_i;
    for (j in 1:p) X_i[, j] = segment(X[, j], pos, n_i);
    for (j in 1:q) Z_i[, j] = segment(Z[, j], pos, n_i);
    vector[n_i] mu_i = X_i * beta;
    matrix[n_i, n_i] Upsilon_i = (sigma * sigma) * diag_matrix(rep_vector(1.0, n_i)) + Z_i * Sigma * transpose(Z_i);
    target += multi_normal_lpdf(Y_i | mu_i, Upsilon_i);
    pos = pos + n_i;
  }
  // population parameters
  target += normal_lpdf(beta | 0, 3);
  target += normal_lpdf(sigma | 0, 3);
  target += normal_lpdf(tau | 0, 3);
  target += lkj_corr_cholesky_lpdf(L | eta);
}
generated quantities {
  corr_matrix[q] Phi = L * transpose(L);
  real rho = Phi[1, 2];
  matrix[q, n] theta;
  int pos;
  pos = 1;
  for (i in 1:n) {
    int n_i = n_is[i];
    vector[n_i] Y_i = segment(Y, pos, n_i);
    matrix[n_i, p] X_i;
    matrix[n_i, q] Z_i;
    for (j in 1:p) X_i[, j] = segment(X[, j], pos, n_i);
    for (j in 1:q) Z_i[, j] = segment(Z[, j], pos, n_i);
    vector[n_i] mu_i = X_i * beta;
    matrix[q, n_i] M = Sigma * transpose(Z_i) * inverse_spd(Z_i * Sigma * transpose(Z_i) + (sigma * sigma) * diag_matrix(rep_vector(1.0, n_i)));
    vector[q] mean_theta_i = M * (Y_i - mu_i);
    matrix[q, q] cov_theta_i = Sigma - M * Z_i * Sigma;
    theta[, i] = multi_normal_rng(mean_theta_i, cov_theta_i);
    pos = pos + n_i;
  }
  vector[n] subject_intercepts = beta[1] + to_vector(theta[1, ]);
  vector[n] subject_slopes = beta[2] + to_vector(theta[2, ]);
}
```

## Glaucoma Disease Progression Data

```{r, eval = FALSE}
glaucoma_longitudinal <- readRDS("glaucoma_longitudinal.rds")
head(glaucoma_longitudinal)
```

```{r, echo = FALSE}
glaucoma_longitudinal <- readRDS("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/data/glaucoma/glaucoma_longitudinal.rds")
head(glaucoma_longitudinal)
```

```{r}
length(unique(glaucoma_longitudinal$eye_id))
nrow(glaucoma_longitudinal)
```

## Fitting the Conditional Model in Stan

```{r, eval = FALSE}
X <- model.matrix(~ time, data = glaucoma_longitudinal)
stan_data <- list(
  N = nrow(glaucoma_longitudinal),
  n = n_eyes,
  p = ncol(X),
  q = ncol(X),
  X = X,
  Z = X,
  Y = glaucoma_longitudinal$mean_deviation,
  Ids = glaucoma_longitudinal$eye_id,
  eta = 1
)
lmm_conditional <- stan_model(model_code = "lmm-conditional.stan")
fit_lmm_conditional <- sampling(lmm_conditional, stan_data, iter = 5000, pars = c("z", "theta"), include = FALSE)
```

```{r, echo = FALSE}
library(rstan)
fit_lmm_conditional <- readRDS("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/r-objects/16-fit-lmm-conditional.rds")
X <- model.matrix(~ time, data = glaucoma_longitudinal)
stan_data <- list(
  N = nrow(glaucoma_longitudinal),
  n = n_eyes,
  p = ncol(X),
  q = ncol(X),
  X = X,
  Z = X,
  Y = glaucoma_longitudinal$mean_deviation,
  Ids = glaucoma_longitudinal$eye_id,
  eta = 1
)
```

## Assessing Convergence

```{r}
traceplot(fit_lmm_conditional, pars = c("beta", "sigma", "tau", "rho"))
```

## Assessing Convergence

```{r}
library(bayesplot)
bayesplot::mcmc_acf(fit_lmm_conditional, regex_pars = c("beta", "sigma", "tau", "rho"))
```

## Posterior Summaries

```{r}
print(fit_lmm_conditional, pars = c("beta", "sigma", "tau", "rho"))
```

## Fitting the Marginal Model in Stan

```{r, eval = FALSE}
X <- model.matrix(~ time, data = glaucoma_longitudinal)
stan_data <- list(
  N = nrow(glaucoma_longitudinal),
  n = n_eyes,
  p = ncol(X),
  q = ncol(X),
  X = X,
  Z = X,
  Y = glaucoma_longitudinal$mean_deviation,
  n_is = as.numeric(table(glaucoma_longitudinal$eye_id)),
  eta = 1
)
lmm_marginal <- stan_model(model_code = "lmm-marginal.stan")
fit_lmm_marginal <- sampling(lmm_marginal, stan_data, iter = 5000)
```

```{r, echo = FALSE}
library(rstan)
fit_lmm_marginal <- readRDS("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/r-objects/16-fit-lmm-marginal.rds")
X <- model.matrix(~ time, data = glaucoma_longitudinal)
stan_data <- list(
  N = nrow(glaucoma_longitudinal),
  n = n_eyes,
  p = ncol(X),
  q = ncol(X),
  X = X,
  Z = X,
  Y = glaucoma_longitudinal$mean_deviation,
  n_is = as.numeric(table(glaucoma_longitudinal$eye_id)),
  eta = 1
)
```

## Assessing Convergence

```{r}
traceplot(fit_lmm_marginal, pars = c("beta", "sigma", "tau", "rho"))
```

## Assessing Convergence

```{r}
library(bayesplot)
bayesplot::mcmc_acf(fit_lmm_marginal, regex_pars = c("beta", "sigma", "tau", "rho"))
```

## Posterior Summaries

```{r}
print(fit_lmm_marginal, pars = c("beta", "sigma", "tau", "rho"))
```

## Comparing LMM versus OLS

```{r}
#| fig-align: "center"
#| fig-height: 8
#| fig-width: 8
#| echo: false

###Fit OLS to each line
n_eyes <- length(unique(glaucoma_longitudinal$eye_id))
ols <- matrix(nrow = n_eyes, ncol = 3)
for (i in 1:n_eyes) {
  dat <- glaucoma_longitudinal[glaucoma_longitudinal$eye_id == i, ]
  mod <- lm(mean_deviation ~ time, data = dat)
  res <- summary(mod)
  pvalue <- pt(coef(res)[, 3], mod$df, lower = TRUE)[2]
  slope <- coef(res)[2, 1]  
  ols[i, ] <- c(i, slope, pvalue)  
}
ols <- data.frame(ols)
colnames(ols) <- c("eye_id", "slope", "p-value")

library(ggplot2)
library(ggExtra)
library(gridExtra)
subject_slopes <- rstan::extract(fit_lmm_marginal, pars = "subject_slopes")$subject_slopes
dat_fig <- data.frame(
  x = ols$slope,
  y = apply(subject_slopes, 2, mean)
)
p <- ggplot(dat_fig, aes(x = x, y = y)) +
  geom_abline(slope = 1, intercept = 0, col = "gray") + 
  geom_point() +  # scatter plot
  theme_minimal() + 
  coord_fixed() +   
  scale_x_continuous(breaks = seq(-4, 1), limits = c(-4, 1.5)) + 
  scale_y_continuous(breaks = seq(-4, 1), limits = c(-4, 1.5)) + 
  labs(x = "OLS Regression", 
       y = "Bayesian Model",
       title = "Eye-Specific Slopes")
ggMarginal(p, type = "histogram")
# p2 <- ggplot(dat_fig, aes(x = x, y = y)) +
#   geom_abline(slope = 1, intercept = 0, col = "gray") + 
#   geom_point() +  # scatter plot
#   theme_minimal() + 
#   coord_fixed() +   
#   scale_x_continuous(breaks = seq(-1, 1, 0.25), limits = c(-1, 1)) + 
#   scale_y_continuous(breaks = seq(-1, 1, 0.25), limits = c(-1, 1)) + 
#   labs(x = "OLS Regression", 
#        y = "Bayesian Model",
#        title = "Zooming into the origin")
# grid.arrange(ggMarginal(p, type = "histogram"), p2, ncol = 2)
```

## Comparing LMM versus OLS

```{r}
#| fig-align: "center"
#| fig-height: 7
#| fig-width: 7
#| echo: false
ggplot(dat_fig, aes(x = x, y = y)) +
  geom_abline(slope = 1, intercept = 0, col = "gray") + 
  geom_point() +  # scatter plot
  theme_minimal() + 
  coord_fixed() +   
  scale_x_continuous(breaks = seq(-1, 1, 0.25), limits = c(-1, 1)) + 
  scale_y_continuous(breaks = seq(-1, 1, 0.25), limits = c(-1, 1)) + 
  labs(x = "OLS Regression", 
       y = "Bayesian Model",
       title = "Zooming into the origin")
```

## Prepare for next class

-   Exam 1 will be released after class!
