---
title: "Gaussian Processes"
author: "Dr. Youngsoo Baek"
date: "2025-03-18"
date-format: "MMM DD, YYYY"
footer: "[🔗 BIOSTAT 725 - Spring 2025](https://biostat725-sp25.netlify.app/)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    html-math-method: mathjax
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(gridExtra)
library(knitr)
library(mvtnorm)
library(coda)
library(tidyr)
library(rstan)
library(bayesplot)

# dataset <- readRDS("glaucoma_longitudinal.rds")
# load("conditional-fit.RData")
# load("marginal-fit.RData")

dataset <- readRDS("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/data/glaucoma/glaucoma_longitudinal.rds")
load("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/r-objects/17-conditional-fit.RData")
load("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/r-objects/17-marginal-fit.RData")
```

## Recalling what went before the break

```{r plot_intro, echo=FALSE}
dataset %>%
  filter(eye_id %in% 1:4) %>% 
  mutate(eye_id = as.factor(eye_id)) %>% 
  ggplot(aes(time, mean_deviation)) + 
  geom_point(aes(colour = eye_id)) +
  geom_line(aes(colour = eye_id), linetype = 2) + 
  labs(x = "Years from baseline",
       y = "Mean Deviation (dB)",
       colour = "Eye")
```

## Is Linear Fit the Right Way? {.small}

There are various ways to fit a model to time-varying behavior of the curves.

```{r plot_smooth, echo=FALSE}
#| fig-width: 7
#| fig-height: 2.7
#| fig-align: "center"
# 1. linear fit
dataset %>%
  filter(eye_id %in% 1:4) %>% 
  mutate(eye_id = as.factor(eye_id)) %>% 
  ggplot(aes(time, mean_deviation)) + 
  geom_point(aes(colour = eye_id)) +
  geom_smooth(aes(colour = eye_id), method = "lm", se= FALSE) + 
  labs(x = "Years from baseline",
       y = "Mean Deviation (dB)",
       colour = "Eye") + ggtitle("Linear Regression on Time")
# 2. LOESS
dataset %>%
  filter(eye_id %in% 1:4) %>% 
  mutate(eye_id = as.factor(eye_id)) %>% 
  ggplot(aes(time, mean_deviation)) + 
  geom_point(aes(colour = eye_id)) +
  geom_smooth(aes(colour = eye_id), method = "loess", se = FALSE) + 
  labs(x = "Years from baseline",
       y = "Mean Deviation (dB)",
       colour = "Eye") + ggtitle("LOESS in time")
```

## A brief agenda

By the end of this lecture you should:

1.  Understand the basic concepts of Gaussian processes (GP) and their usefulness in statistical modeling.

2.  Compute posterior and predictive inference for a simple GP model.

3.  Use GP as a building block in modeling time-correlated data.

## Motivation {.midi}

Let $i$ index distinct eyes ($i=1,2,\ldots,n$) and $t$ index within-eye data in time ($t=1,2,\ldots,n_i$).

Suppose we are interested in effects of time-invariant predictors (`age` and `iop`):

$$Y_{it} = \underbrace{\mathbf{x}_i\boldsymbol{\beta}}_{\text{Fixed in time}} + \underbrace{\eta_{it}}_{\text{Varying in time}} + \epsilon_{it}$$

In a previous lecture, we included only `time` ($X_{it}$) and fit a random slope regression.

$$\eta_{it} = \theta_{0i} + \underbrace{X_{it}}_{=Time}\theta_{1i}.$$

What if we want a **nonlinear model** for time effects?

## Exploratory Analysis of Population Predictors

```{r eda_vis, echo=FALSE}
dataset_binned <- dataset %>%
 mutate(age_group = cut(age, breaks = c(0, 50, 60, 70, 100)),
        iop_group = cut(iop, breaks = c(0, 10, 15, 20, 30)))
ggplot(dataset_binned, aes(time, mean_deviation)) + 
  geom_line(aes(group = eye_id), alpha = .4) + 
  facet_wrap(~ age_group) + ggtitle("Binned by age") + 
  labs(x = "Years from baseline",
       y = "Mean Deviation (dB)")
```

## Exploratory Analysis of Population Predictors

```{r, echo=FALSE}
ggplot(dataset_binned, aes(time, mean_deviation)) + 
  geom_line(aes(group = eye_id), alpha = .4) + 
  facet_wrap(~ iop_group) + ggtitle("Binned by iop (mm Hg)") + 
  labs(x = "Years from baseline",
       y = "Mean Deviation (dB)")
```

## A review of computation / linear algebra needed to sample {.small}

How can we sample a 3D normal random vector?

```{r toy_code, eval=FALSE}
p <- 3
mu <- c(3, -1, 5)
Sigma <- matrix(
  c(
    1.0, 0.6, 0.3,
    0.6, 1.0, 0.6,
    0.3, 0.6, 1.0
  ),
  nrow = 3, ncol = 3
)
z <- rnorm(p)
R <- t(chol(Sigma)) ## R %*% t(R) == Sigma
x <- R %*% z

# This is the same as
x <- rmvnorm(1, mu, Sigma)
```

From previous lecture: `R` decomposes into `D` (diagonal) and `L` (Cholesky factor of a correlation matrix).

## Constructing a Gaussian Random Function in Time {.midi}

What about "random functions"? Can we extend this idea to arbitrarily many time points? What does that even mean?

1.  We start at time $t=0$ from $X_0 = 0$.

2.  When time passes by amount of $h > 0$, we want

$$X_{t+h} - X_t \sim N(0,h).$$ Note: $\mathbb{E}[X_t^2] = \mathbb{V}(X_t) = t$ (Why?)

3.  Each increment will be independent from each other.

Now think of *passing to the limit*: Somehow it works (!) and we have a process defined **at every** $t$.

## Pointwise Observations from a Brownian Motion {.small}

Suppose we observe this process $X_t$ at time points $(t_1,t_2,\ldots,t_N)$. Then the distribution of the following vector is a multivariate normal:

$$
\begin{bmatrix}
X_1\\ \vdots \\ X_N
\end{bmatrix}\sim N\left(\mathbf{0},\boldsymbol{\Sigma}\right),\; \boldsymbol{\Sigma} = \begin{bmatrix}
t_1 & t_1 & \cdots & t_1 \\
t_1 & t_2 & \cdots & t_2 \\
\vdots & \vdots & \ddots & \vdots \\
t_1 & t_2 & \cdots & t_N
\end{bmatrix}
$$

The covariance structure comes from solving the equation:

$$
h = \mathbb{V}(X_{t+h} - X_t) = \underbrace{\mathbb{E}[X_{t+h}^2]}_{=t+h} + \underbrace{\mathbb{E}[X_t^2]}_{=t} - 2\mathbb{C}(X_{t+h},X_t).
$$

## Visualizing the Brownian Motion

```{r brownian_motion, echo=FALSE}
#| fig-align: "center"
time_grid <- seq(0, 5, length.out = 1000)
h <- diff(time_grid)
incr <- rnorm(length(h), 0, h)
x <- cumsum(c(0, incr))
data.frame(y = x, x = time_grid) |>
  ggplot(aes(x = x, y = y)) + 
  geom_line() + 
  labs(x = "Time",
       y = "X",
       title = "Brownian Motion")
```

## What if I want different correlation structures? {.midi}

For longitudinal data,

-   We are not interested in an indefinitely long time span.

-   Within the time window, the data can exhibit stable behavior.

i.  *Hypothesis 1* : The marginal variability of $X_t$ should stay the same.

ii. *Hypothesis 2* : Time correlation should decay in, and **only** depend on, the amount of time elapsed.

iii. *Hypothesis 3* : We have expectations about the span of correlation and smoothness of the process.

These are all **a priori** hypotheses. The data may come from something very different!

## Stationary Gaussian processes

A different process results from a **stationary kernel**:

$$
\mathbb{C}(X_t, X_s) = C(|t-s|).
$$

We want $C(0) = \sigma^2 > 0$ and $C\to 0$ as $h = |t-s|\to\infty$. Some common choices:

-   Exponential kernel: $C(h) = \sigma^2\exp(-h/\rho)$

-   Square exponential kernel: $C(h) = \sigma^2\exp\{-h^2/(2\rho^2)\}$

-   Matérn kernels

## Visualizing the process

```{r exp_kernel, echo=FALSE}
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 2.5
time_grid <- seq(0, 5, length.out = 50)
kernel <- exp(-.5 / (.5) * as.matrix(dist(time_grid)))
root <- chol(kernel + diag(1e-8, nrow(kernel)))
x <- t(root) %*% rnorm(50)
data.frame(y = x, x = time_grid) |>
  ggplot(aes(x = x, y = y)) + 
  geom_line() + 
  labs(x = "Time",
       y = "X",
       title = "Exponential Processs")
```

```{r sqexp_kernel, echo=FALSE}
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 2.5
time_grid <- seq(0, 5, length.out = 50)
kernel <- exp(-.5 / (.5)^2 * as.matrix(dist(time_grid))^2)
root <- chol(kernel + diag(1e-8, nrow(kernel)))
x <- t(root) %*% rnorm(50)
data.frame(y = x, x = time_grid) |>
  ggplot(aes(x = x, y = y)) + 
  geom_line() + 
  labs(x = "Time",
       y = "X",
       title = "Square Exponential Processs")
```

## Interpreting Hyperparameters and Kernels

Often, we fix the overall **kernel function** while learning its "bandwidth" $\rho$ due to poor identifiability from the data.

```{r kernel_vis, echo = FALSE}
#| fig-align: "center"
#| fig-height: 6
#| fig-width: 12
rho_grid <- seq(0.5, 2, by = 0.2)
dist_grid <- seq(0, 5, by = 0.05)
cov_grid <- expand_grid(rho = rho_grid, d = dist_grid) %>% 
  mutate(exponential = exp(-d / rho), square_exp = exp(-.5 * d^2 / rho^2)) %>% 
  pivot_longer(c("exponential", "square_exp"), names_to = "kernel", values_to = "cor")
cov_grid$kernel <- as.factor(cov_grid$kernel)
levels(cov_grid$kernel) <- c("Exponential Kernel", "Squared Exponential Kernel")
cov_grid$rho <- as.factor(cov_grid$rho)
ggplot(cov_grid) +
  geom_line(aes(d, cor, group = rho, colour = rho)) + 
  facet_wrap(~kernel) + 
  labs(x = "Distance in time (h)", 
       y = "Correlation",
       colour = expression(rho))
```

## Default Prior Choices {.midi}

-   Stan team often recommends $\rho^{-1} \sim Gamma(5,5)$

-   *A priori* the prior is concentrated around 1. For a prior mean, distance increase of 1 corrresponds to a multiplicative decay of correlation by $e^{-1}\sim 37\%$.

-   Beware of the units! A **year** correlation is very different from that over a **day**.

```{r prior_vis, echo = FALSE}
#| fig-align: "center"
#| fig-height: 5
#| fig-width: 8
grid <- seq(0.1, 5, length.out = 100)
dens <- dgamma(grid, 5, 5)
ggplot(cbind.data.frame(x = grid, y = dens)) + 
  geom_line(aes(x = x, y = y), lwd = 1.5) + 
  geom_vline(aes(xintercept = 1), col = "red", lwd = 1.5) + 
  xlab(expression(1/rho)) + ylab("Prior Density")
```

## Gaussian Process Covariance Functions in Stan

A list of available kernels is available [here](https://mc-stan.org/docs/functions-reference/matrix_operations.html#gaussian-process-covariance-functions).

The squared exponential kernel (exponentiated quadratic kernel) is given by:

$$C(\mathbf{x}_i,\mathbf{x}_j) = \sigma^2 \exp\left(-\frac{|\mathbf{x}_i - \mathbf{x}_j|^2}{2l^2}\right)$$

```{stan output.var = "kernel", eval = FALSE}
matrix gp_exp_quad_cov(array[] real x, real sigma, real length_scale)
```

For us, $\mathbf{x}$ is 1D (time)...but it does not have to be!

## Prior Sampling in Stan {.midi}

Once we have a GP model, sampling from the prior is equivalent to sampling from a multivariate normal.

```{stan output.var = "prior_sampling", eval = FALSE}
data {
  int<lower=1> N;
  array[N] real x;
  real<lower=0> sigma;
  real<lower=0> l;
}
transformed data {
  // 0. A "small nugget" to stabilize matrix root computation
  // This is for.better numerical stability in taking large matrix roots
  real delta = 1e-9;

  // 1. Compute the squared exponential kernel matrix
  // x is the time variable
  vector[N] mu = rep_vector(0, N);
  matrix[N, N] R_C;
  matrix[N, N] C = gp_exp_quad_cov(x, sigma, l);
  for (i in 1:N) {
    C[i, i] = C[i, i] + delta;
  }

  // 2. Compute the root of C by Cholesky decomposition
  R_C = cholesky_decompose(C);
}
generated quantities {
  // 3. Sample from the prior: multivariate_normal(0, C)
  f ~ multi_normal_cholesky(mu, R_C)
}
```

## Posterior inference {.midi}

For each $i$-th eye: time effect vector is given an independent prior based on the GP model.

$$\boldsymbol{\eta}_i = (\eta_{i1},\ldots,\eta_{in_i})^\top \stackrel{ind}{\sim} N_{n_i}(\mathbf{0},\mathbf{C}_i)$$

Each $\eta_{it}$ is a pointwise value of the process $\eta_i(t) = \eta_{it}$.

$$\mathbf{C}_i = \begin{bmatrix}
C(0) & C(|t_{i1}-t_{i2}|) & \cdots & C(|t_{i1} - t_{in_i}|)\\
C(|t_{i1} - t_{i2}|) & C(0) & \cdots & C(|t_{i,2} - t_{in_i}|)\\
\vdots & \vdots & \ddots & \vdots
\end{bmatrix}$$

**Remember**: Different eyes need not have the same number of encounters, nor need they have measurements at the same time.

## The full model {.midi}

For $i$ ($i = 1,\ldots,n$) and $t$ ($t = 1,\ldots,n_i$),

\begin{align*}
Y_{it} &= \mathbf{x}_{i}\boldsymbol{\beta} + \eta_{i}(t) + \epsilon_{it}, \quad \epsilon_{it} \stackrel{iid}{\sim} N(0,\sigma^2),\\
\boldsymbol{\eta}_i &\stackrel{ind}{\sim} N_{n_i}(\mathbf{0},\mathbf{C}_i),\\
\boldsymbol{\Omega} &\sim f(\boldsymbol{\Omega}).
\end{align*}

-   $\boldsymbol{\Omega}$ represents the population parameters: $\boldsymbol{\beta}$, $\sigma^2$, and any kernel parameters determining $\mathbf{C}_i$

-   For concreteness: we use a squared exponential kernel $C$ and model $\mathbb{C}(\eta_i(t),\eta_i(t')) = \alpha^2\exp(-|t-t'|^2/2\rho^2).$ Thus $\boldsymbol{\Omega} = (\boldsymbol{\beta},\sigma^2,\alpha,\rho)$.

## First Look at the Data

```{r data}
head(dataset)
```

## Model Specification {.midi}

We want to fit a model estimating the effects of age and iop on mean deviation, adjusting for nonlinear time effects:

\begin{align*}
Y_{it} &= \underbrace{\beta_0 + \text{age}_i\beta_1 + \text{iop}_i\beta_2}_{=\mathbf{x}_{i}\boldsymbol{\beta}} + \eta_{it} + \epsilon_{it}\\
\boldsymbol{\eta}_i &= \begin{bmatrix} \eta_{i1}\\ \vdots\\ \eta_{in_i} \end{bmatrix} \stackrel{iid}{\sim} N(0,\mathbf{C}_i)\\
\epsilon_{it} &\stackrel{iid}{\sim} N(0,\sigma^2)
\end{align*}

Our model for $\eta_{it}$ is *nonlinear* in time: `time` enters the covariance structure of $\mathbf{C}_i$ producing a function $\eta_i(t) = \eta_{it}$.

## Preparing the Data

A few data processing steps are needed to handle the data structure with more ease.

-   $N$ will denote the number of **all** observations: $N = \sum_{i=1}^{n} n_i$.

-   $n$ will denote the total number of **eyes**.

-   $p$ will denote the number of predictors fixed across time ($p=3$: intercept, slopes for `age` and `iop`).

-   A separate vector of the number of observations ($n_i$) for each eye will be stored as `s`.

## Preparing the Data {.midi}

```{r data_proc, message=FALSE, warning=FALSE}
library(dplyr)

# Group predictors fixed across time
fixed_df <- dataset %>% 
  group_by(eye_id) %>% 
  reframe(age = unique(age), iop = unique(iop))
Xmat <- model.matrix(~age + iop, data = fixed_df)

# Number of measurements for each eye
groupsizes <- dataset %>% 
  group_by(eye_id) %>% 
  summarise(n = n()) %>% 
  pull(n)

stan_data <- list(
  N = dim(dataset)[1],
  n = max(dataset$eye_id),
  p = dim(Xmat)[2],
  t = dataset$time,
  Y = dataset$mean_deviation,
  s = groupsizes,
  X = Xmat
)
```

## Conditional GP Model {.midi}

```{stan output.var = "conditional_fit", eval = FALSE}
data {
  int<lower=0> N;       // total number of observations
  int<lower=1> n;       // number of eyes
  int<lower=1> p;       // fixed effects dimension
  vector[N] Y;          // observation
  matrix[n, p] X;        // fixed effects predictors
  array[N] real t;      // obs time points
  array[n] int s;       // sizes of within-pt obs
}
transformed data {
  real delta = 1e-9;
}
parameters {
  // Fixed effects model
  vector[p] beta;
  real<lower=0> sigma;
  // GP parameters
  vector[N] eta;
  real<lower=0> alpha;
  real<lower=0> rho;
}
transformed parameters {
  vector[n] mu = X * beta;
}
model {
  beta ~ normal(0,3);
  z ~ std_normal();
  alpha ~ std_normal();
  sigma ~ std_normal();
  rho ~ inv_gamma(5,5);

  vector[N] mu_rep;
  vector[N] eta;
  
  int pos;
  pos = 1;
  // Ragged loop computing the mean for each time obs
  for (i in 1:n) {
    // GP covariance for the k-th eye
    int n_i = s[i];
    int pos_end = pos + n_i - 1;
    matrix[n_i, n_i] R_C;
    matrix[n_i, n_i] C = gp_exp_quad_cov(segment(t, pos, n_i), alpha, rho);
    for (j in 1:n_i) {
      // adding a small term to the diagonal entries
      C[j, j] = C[j, j] + delta;
    }
    R_C = cholesky_decompose(C);
    
    // Mean of data at each time
    mu_rep[pos:pos_end] = rep_vector(mu[i], n_i);
    // GP for the i-th eye
    eta[pos:pos_end] = R_C * segment(z, pos, n_i);
    pos = pos_end + 1;
  }
  // Normal observation model centered at mu_rep + eta
  Y ~ normal(mu_rep + eta, sigma);
}
```

## Assessing Convergence

```{r traceplot1}
traceplot(gp_condl_fit, pars = c("beta", "sigma", "alpha", "rho"))
```

## Assessing Convergence

```{r acf1}
bayesplot::mcmc_acf(gp_condl_fit, regex_pars = c("beta", "sigma", "alpha", "rho"))
```

## Marginal Model Specification {.midi}

The conditional model is pretty slow (*can be improved*) and results in poor mixing (*can be run longer*). Marginalizing the random effect can stabilize the computation.

In vector notation, our observation model for the $i$-th eye is

$$
\mathbf{Y}_i = \begin{bmatrix} Y_{i1}\\ \vdots \\ Y_{in_i} \end{bmatrix}
= \mathbf{x}_i\boldsymbol{\beta} + \boldsymbol{\eta}_i + \boldsymbol{\epsilon}_i
$$ By marginalizing out $\boldsymbol{\eta}_i$, we obtain

$$
\mathbf{Y}_i \sim N_{n_i}(\mathbf{x}_i\boldsymbol{\beta},\mathbf{C}_i + \sigma^2\mathbf{I}_{n_i})
$$

## Fitting a Marginal Model {.midi}

Only the `model` segment has to be changed.

```{stan output.var = "marginal", eval = FALSE}
model {
  beta ~ normal(0,3);
  alpha ~ std_normal();
  sigma ~ std_normal();
  rho ~ inv_gamma(5,5);

  int pos;
  pos = 1;
  // Ragged loop computing joint likelihood for each eye
  for (i in 1:n) {
    // GP covariance for the k-th eye
    int n_i = s[i];
    vector[n_i] mu_rep;
    
    matrix[n_i, n_i] R_C;
    matrix[n_i, n_i] C = gp_exp_quad_cov(segment(t, pos, n_i), alpha, rho);
    for (j in 1:n_i) {
      // Add noise variance to the diagonal entries
      C[j, j] = C[j, j] + square(sigma);
    }
    R_C = cholesky_decompose(C);
    // Marginal model for the i-th eye
    mu_rep = rep_vector(mu[i], n_i);
    target += multi_normal_cholesky_lpdf(to_vector(segment(Y, pos, n_i)) | mu_rep, R_C);
    pos = pos + n_i;
  }
}
```

## Assessing Convergence

```{r traceplot2}
traceplot(gp_marginal_fit, pars = c("beta", "sigma", "alpha", "rho"))
```

## Assessing Convergence

```{r acf2}
bayesplot::mcmc_acf(gp_marginal_fit, regex_pars = c("beta", "sigma", "alpha", "rho"))
```

## Estimates of eye-level Predictors

```{r ols_comparison, echo = FALSE, message = FALSE}
bayesplot::mcmc_areas(gp_marginal_fit, pars = c("beta[2]", "beta[3]"), prob = 0.95) +
  scale_y_discrete(labels = c("beta[3]" = "IOP (mm Hg)", "beta[2]" = "Age (years)")) +
  ggtitle("Coefficient Posterior Densities (95% credible intervals)")
```

## Learned GP hyperparameter {.midi}

-   The posterior mean of $\rho$ (in year) is very large: the pattern of mean deviation change is very gradual over time.

-   *A posteriori* we believe mean deviations of an eye, 10 years apart and adjusted for age and iop, are still strongly correlated ($e^{-5/6}\sim 43\%$).

```{r post_vis, echo = FALSE}
#| fig-align: "center"
bayesplot::mcmc_hist(gp_marginal_fit, pars = "rho") + 
  xlab(expression(rho)) + 
  scale_x_continuous(limits = c(0, 15)) + 
  geom_vline(aes(xintercept = 1), colour = 2) + 
  ggtitle("Posterior against Prior Mean")
```

## Random effects covariance structure for the first 5 years

```{r cov_heatmap, echo = FALSE}
stan_data <- list(
  N = dim(dataset)[1],
  n = max(dataset$eye_id),
  p = dim(Xmat)[2],
  Y = dataset$mean_deviation,
  t = dataset$time,
  s = groupsizes,
  X = Xmat,
  t_pred = seq(0, 5, by = 0.25),
  N_pred = 21
)

t_pred <- stan_data$t_pred

Cp <- rstan::extract(gp_marginal_fit)$Cp
Cp_mean <- as.data.frame(cov2cor(apply(Cp, 2:3, mean)))
Cp_tbl <- tidyr::gather(Cp_mean)
Cp_tbl$x <- rep(t_pred, length(t_pred))
Cp_tbl$y <- rep(t_pred, each = length(t_pred))
ggplot(data = Cp_tbl) + geom_tile(aes(x, y, fill = value)) + 
  coord_fixed() +
  labs(x = "Years from baseline", y = "Years from Baseline", fill = "Correlation") + 
  ggtitle("Posterior Mean Correlation Matrix") +
  theme_classic()
```

## Predictive inference

The model is *continuous* in nature and defined, in principle, for all times. This is very convenient for visualiztion, interpolation, and forecasting.

$$
\mathbf{Y}_i = \underbrace{\mathbf{f}_{i}}_{ = \mathbf{x}_{i}\boldsymbol{\beta} + \boldsymbol{\eta}_{i} } + \boldsymbol{\epsilon}_i
$$

The vector $\mathbf{f}_{i}$ may be interpreted as the **denoised latent process** of an eye-specific mean deviation over time: $(f_{i1},\ldots,f_{in_i})^\top$.

$$
\mathbb{E}[f_{it}] = \mathbf{x}_i\boldsymbol{\beta},\; \mathbb{C}(f_{it}, f_{it'}) = \mathbb{C}(\eta_{it},\eta_{it'})
$$

## Predictive Inference using Normal Conditioning {.small}

Say $\mathbf{f}^{pred}_i$ is the values of $f_i$ at $m_i$ **new time points** which we want to predict, based on our GP model:

\begin{align*}
\mathbf{f}^{pred}_i &\sim N_{m_i}(\mathbf{x}_i\boldsymbol{\beta},\mathbf{C}^{pred}_i)\\
\mathbf{Y}_i = \mathbf{f}_i + \boldsymbol{\epsilon}_i &\sim N_{n_1}(\mathbf{x}_i\boldsymbol{\beta},\mathbf{C}_i + \sigma^2\mathbf{I}_{n_i})
\end{align*}

An analytical formula exists for the **conditional distribution** of $\mathbf{f}^{pred}$ given observed $Y_{it}$ at $n_i$ time points.

$$
\mathbf{f}^{pred}_i | \mathbf{Y}_i \sim N\left(\mathbf{x}_i\boldsymbol{\beta} + \mathbf{k}_i(\mathbf{C}_i + \sigma^2\mathbf{I})^{-1}(\mathbf{Y}_i - \mathbf{x}_i\boldsymbol{\beta}), \mathbf{C}_i^{pred}-\mathbf{k}_i(\mathbf{C}_i + \sigma^2\mathbf{I})^{-1}\mathbf{k}_i^\top\right)
$$ where $\mathbf{k}_i = \mathbb{C}(\mathbf{f}^{pred}_i, \mathbf{f}_i)$ ("cross-covariances").

This stems from a more general fact about conditional distributions of jointly normal vectors: e.g., $\mathbf{f}_i^{pred}$ and $\mathbf{f}_i$ need not have the same mean.

## New Data for Predictive Inference

Suppose we are interested in understanding the mean deviation trend for the first 5 years from baseline ($t^{pred}\in [0,5]$).

```{r new_data, eval = FALSE}
stan_data <- list(
  N = dim(dataset)[1],
  n = max(dataset$eye_id),
  p = dim(Xmat)[2],
  Y = dataset$mean_deviation,
  t = dataset$time,
  s = groupsizes,
  X = Xmat,
  # Time window for which we want predictive
  t_pred = seq(0, 5, by = 0.25),
  # Total length of this window
  N_pred = 21
)
```

## Stan Implementation {.small}

See the [Stan Help page](https://mc-stan.org/docs/stan-users-guide/gaussian-processes.html) for details.

```{stan output.var = "predictive_inference", eval = FALSE}
functions {
  // Analytical formula for latent GP conditional on Gaussian observations
  vector gp_pred_rng(array[] real x_pred,
                     vector Y,
                     array[] real x,
                     real mu,
                     real alpha,
                     real rho,
                     real sigma,
                     real delta) {
    int N1 = rows(Y);
    int N2 = size(x_pred);
    vector[N2] f_pred;
    {
      matrix[N1, N1] L_Sigma;
      vector[N1] Sigma_div_y;
      matrix[N1, N2] C_x_xpred;
      matrix[N1, N2] v_pred;
      vector[N2] fpred_mu;
      matrix[N2, N2] cov_fpred;
      matrix[N2, N2] diag_delta;
      matrix[N1, N1] Sigma;
      Sigma = gp_exp_quad_cov(x, alpha, rho);
      for (n in 1:N1) {
        Sigma[n, n] = Sigma[n, n] + square(sigma);
      }
      L_Sigma = cholesky_decompose(Sigma);
      Sigma_div_y = mdivide_left_tri_low(L_Sigma, Y - mu);
      Sigma_div_y = mdivide_right_tri_low(Sigma_div_y', L_Sigma)';
      C_x_xpred = gp_exp_quad_cov(x, x_pred, alpha, rho);
      fpred_mu = (C_x_xpred' * Sigma_div_y);
      v_pred = mdivide_left_tri_low(L_Sigma, C_x_xpred);
      cov_fpred = gp_exp_quad_cov(x_pred, alpha, rho) - v_pred' * v_pred;
      diag_delta = diag_matrix(rep_vector(delta, N2));

      f_pred = multi_normal_rng(fpred_mu, cov_fpred + diag_delta);
    }
    return f_pred;
  }
}
//...
generated quantities {
  matrix[n,Np] f_pred;
  matrix[Np,Np] Cp;
  Cp = gp_exp_quad_cov(t_pred, alpha, rho);
  // Posterior predictive on fixed time grid for all eyes
  int pos;
  pos = 1;
  for (i in 1:n) {
    int n_i = s[i];
    f_pred[i,] = mu[i] + 
      gp_pred_rng(
          t_pred,
          segment(Y, pos, n_i),
          segment(t, pos, n_i),
          mu[i], 
          alpha,
          rho,
          sigma,
          delta
        )';
    pos = pos + n_i;
  }
}
```

## Visualizing latent effects using predictive formulae

```{r predictive, echo = FALSE}
t_pred <- stan_data$t_pred
# Select a few eyes
f_pred <- rstan::extract(gp_marginal_fit)$fp[,1:6,]
mean_pred <- apply(f_pred, 2:3, mean)
low_pred <- apply(f_pred, 2:3, quantile, probs = 0.025)
hi_pred <- apply(f_pred, 2:3, quantile, probs = 0.975)
plot_df <- data.frame(mean_pred) %>% 
  tidyr::gather() %>% 
  mutate(id = rep(1:6, length(t_pred)), 
         t = rep(t_pred, each = 6), 
         type = "fit",
         low = c(low_pred), 
         hi = c(hi_pred)) %>% 
  select(-key) %>% 
  bind_rows(
    dataset %>% filter(eye_id <= 6, time <= 5) %>% select(mean_deviation, time, eye_id) %>% 
      rename(value = mean_deviation, id = eye_id, t = time) %>% 
      mutate(type = "data")
  )
ggplot() + 
  geom_point(data = plot_df %>% filter(type == "data"), aes(t, value)) + 
  geom_line(data = plot_df %>% filter(type == "fit"), aes(t, value)) + 
  geom_ribbon(data = plot_df %>% filter(type == "fit"), aes(t, ymin = low, ymax = hi), fill="blue", alpha=0.25) + 
  facet_wrap(~id, scales = "free_y") + 
  xlab("Years from baseline") + 
  ylab("Mean Deviation (dB)") + 
  ggtitle("Latent Mean Posterior (95% pointwise intervals)")
```

## Stan speed concerns

-   GP is notorious for not being scalable. Tl;dr is the need for matrix root / inverse computation.

-   For datasets covered in this class, Stan works well. For research purposes, worth exploring other specialized toolkits.

i.  Code MCMC yourself to make it faster (e.g., using `Rcpp`)

ii. Do a bit of software research on [Wikipedia](https://en.wikipedia.org/wiki/Comparison_of_Gaussian_process_software)

iii. Scalable Approximations (some of them will be coming soon!)

## A summary

-   GP is a flexible, high-dimensional model for handling correlated measurements over time.

-   With some basic knowledge about conditioning Gaussian random variables, we can implement posterior computation and prediction / interpolation.

-   Programming in Stan is straightforward but can be expensive with large number of observations.

## Prepare for next class

-   Work on [HW 04](https://biostat725-sp25.netlify.app/hw/hw-04) which is due March 25

-   Complete reading to prepare for next Thursday's lecture

-   Thursday's lecture: Using GP to model spatial data
