---
title: "Disease Mapping"
author: "Prof. Sam Berchuck"
date: "2025-03-25"
date-format: "MMM DD, YYYY"
footer: "[🔗 BIOSTAT 725 - Spring 2025](https://biostat725-sp25.netlify.app/)"
logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    html-math-method: mathjax
filters:
  - parse-latex
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
bibliography: references.bib
---

```{r, echo=FALSE, message=FALSE, warning=FALSE, }
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(knitr)
library(bayesplot)
library(rstan)
library(igraph)
library(tidycensus)
library(sf)
library(rnaturalearth)
library(tidyverse)
library(tigris)
library(scales)
library(spdep)
library(reshape2)
nc_counties_sf <- st_read("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/data/covid/nc_counties.shp", quiet = TRUE)
```

## Review of last lecture

-   During our last lecture, we learned about Gaussian processes.

-   We learned how to apply Gaussian processes to longitudinal (or time-series) data.

-   The longitudinal setting is one-dimensional (i.e., time). Today we will learn about applying Gaussian processes in two-dimensions (i.e. space).

## Goal of an Areal Data Spatial Analysis

- Account for noise due to spatial variability in order to provide smoothed estimates across space

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
#| layout-ncol: 1
ggplot(data = nc_counties_sf) +
  geom_sf(fill = "lightblue", color = "black") +
  geom_sf(aes(fill = DEATHS), shape = 16, size = 2) +
  theme_minimal() +
  labs(
    title = "2020 COVID Deaths by County",
    fill = "Deaths"
  ) +
  scale_fill_viridis_c() +  # Custom labels for the breaks
  coord_sf()  # Ensures the map is properly projected
```

## Motivating Data



## Spatial Correlation: Areal Data

- How to induce spatial correlation between areal units? 
	
  - Distances between centroids (possibly population weighted); may be inappropriate for oddly shaped regions of varying sizes (great for equal sized grid though).

  - Neighborhood structure of your spatial region; are two regions neighbors?
  
- Correlation introduced through spatial random effects.

## Bayesian Model Fitting

- CAR models are preferred in the Bayesian setting due to the conditional definition of the model

- $\theta_i$ parameters are typically updated individually.

- Depending on the likelihood, conjugacy may be available, useful for Gibbs sampling.

- Today, CAR models can be implemented using probabilistic programming languages (e.g., Stan).

## Inducing Spatial Dependency

The most common approach for inducing spatial dependency in an areal data setting using the intrinsic conditional autoregressive (ICAR) process,

$$\boldsymbol{\theta} | \tau^2 \sim \text{ICAR}\left(\tau^2\right)$$

$$\theta_{i} | \boldsymbol{\theta}_{-i}, \tau^2 \sim \mathcal N \left({\frac{\sum_{j=1}^n w_{ij}\theta_{j}}{\sum_{j=1}^n w_{ij}}},\frac{\tau^2}{\sum_{j=1}^n w_{ij}}\right)$$

- $\boldsymbol\theta_{-j}$: Vector of $\theta_{i}$ parameters with $\theta_{j}$ removed

- $\boldsymbol{\beta}_0 | \tau_0^2 \sim \text{ICAR}\left(\tau_0^2\right),\quad \boldsymbol{\beta}_1 | \tau_1^2 \sim \text{ICAR}\left(\tau_1^2\right)$

- Not a proper prior distribution (implications for MCMC algorithms!)

## A Proper CAR Process

$$\boldsymbol{\theta} = \left\{\theta_1, \ldots, \theta_n\right\}^{\text{T}};\  \boldsymbol{\theta} | \rho, \tau^2 \sim \text{Leroux CAR}\left(\rho,\tau^2\right)$$

$$\theta_i | \boldsymbol{\theta}_{-i}, \rho, \tau^2 \sim \mathcal N \left({\frac{\rho \sum_{j=1}^n w_{ij} \theta_j}{\rho \sum_{j=1}^n w_{ij} + 1 - \rho}}, \frac{\tau^2}{\rho \sum_{j=1}^n w_{ij} + 1 - \rho}\right)$$

- Proper for $\rho \in [0,1)$

- $\rho=1$ gives us the ICAR model

  - $\rho$ is given a prior distribution and estimated by the data
  
  - $\text{COR}(\theta_k,\theta_j|\boldsymbol{\theta}_{-kj}) = \frac{\rho w_{kj}}{\sqrt{(\rho \sum_{i=1}^n w_{ki} + 1 - \rho)(\rho \sum_{i=1}^n w_{ji} + 1-\rho)}}$

## Prepare for next class

-   Work on HW 04, which is due before class on Tuesday.

-   Complete reading to prepare for next Tuesday's lecture

-   Tuesday's lecture: Disease mapping
