---
title: "Scalable Gaussian Processes #1"
author: "Christine Shen"
date: "2025-04-01"
date-format: "MMM DD, YYYY"
footer: "[🔗 BIOSTAT 725 - Spring 2025](https://biostat725-sp25.netlify.app/)"
# logo: "../images/logo.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: false
    transition: fade
    slide-number: true
    incremental: false 
    chalkboard: true
    html-math-method: mathjax
execute:
  freeze: auto
  echo: true
knitr:
  opts_chunk: 
    R.options:      
    width: 200
# bibliography: ../../doc/reference.bib
editor_options: 
  chunk_output_type: console
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(sf)
library(rnaturalearth)
```


## Review of the last lecture

TBU

## Motivating dataset {.midi}

Recall we worked with a dataset on women aged 15-49 sampled from the 2013-14 Democratic Republic of Congo (DRC) Demographic and Health Survey. 

- There are ~8,600 women nested in ~500 survey clusters. 

- Variables are:

  - `loc_id`: location id (i.e. survey cluster).

  - `hemoglobin`: hemoglobin level (g/dL).

  - `anemia`: anemia classifications.
  
  - `age`: age in years.

  - `urban`: urban vs. rural.

  - `LATNUM`: latitude.

  - `LONGNUM`: longitude.

## Motivating dataset {.midi}

```{r, echo = FALSE}
data <- read.csv(file = "/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/data/drc/hemoglobin_anemia.csv")
data <- data[complete.cases(data),-1]

# Convert to spatial dataset and merge DRC data
data_sf <- st_as_sf(data, coords = c("LONGNUM", "LATNUM"), crs = 4326)
congo_states_map <- ne_states(country = "Democratic Republic of the Congo", returnclass = "sf") %>%
  select(name,geometry)
# Ensure that the CRS for the country and grid points match
congo_states_map <- st_transform(congo_states_map, crs = 4326) 
data_sf_drc <- st_intersection(data_sf, congo_states_map)

head(data)
```

::: callout-important
## Modeling goals:

  - Learn the associations between age and urbanality and hemoglobin, accounting for unmeasured spatial confounders.
  
  - Create a predicted map of hemoglobin across the spatial surface controlling for age and urbanality, with uncertainty quantification.
  
:::

## Map of Sud-Kivu state {.midi}

Last time, we focused on one state with ~500 women at ~30 locations.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

states_of_interest <- c("Sud-Kivu")
data_sf_states_of_interest <- data_sf_drc %>% 
  filter(name %in% states_of_interest) %>%
  group_by(loc_id, urban) %>%
  summarise(hemoglobin=mean(hemoglobin))

ggplot() +
  geom_sf(data = subset(congo_states_map, name %in% states_of_interest), fill = "lightblue", color = "black", size = 0.2) +
  # Plot the points from your data
  geom_sf(data = data_sf_states_of_interest, aes(color = hemoglobin), shape = 16, size = 2) +
  scale_color_viridis_b() + 
  theme_minimal() +
  labs(title = "Average hemoglobin at each community",
       subtitle = paste("State:", paste(states_of_interest, collapse = ", ")),
       color = "Hemoglobin (g/dL)") 
```

## Map of the DRC{.midi}

Today we will extend the analysis to the full dataset.

```{r}
#| echo: false
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 8
#| fig-height: 8

data_mean <- data %>%
  group_by(loc_id, urban, LATNUM, LONGNUM) %>%
  summarise(hemoglobin=mean(hemoglobin))
data_sf_whole <- st_as_sf(data_mean, coords = c("LONGNUM", "LATNUM"), crs = 4326)

ggplot() +
  # Plot the map of DRC
  geom_sf(data = congo_states_map, fill = "lightblue", color = "black") +
  geom_sf(data = subset(congo_states_map, name %in% states_of_interest), 
          fill = NA, color = "#a50f15", linewidth = 1.5) +
  geom_sf(data = data_sf_whole, aes(color = hemoglobin), shape = 16, size = 2) +
  scale_color_viridis_b() + 
  # Customize the plot appearance
  theme_minimal() +
  labs(title = "Average Hemoglobin Across Communities in the Democratic Republic of Congo",
       subtitle = "Points represent community locations with hemoglobin averages",
       x = "Longitude", y = "Latitude", color = "Hemoglobin (g/dL)") 

```

## Modeling{.midi}

\begin{align*}
  Y_j(\mathbf{u}_i) &= \alpha + \mathbf{x}_j(\mathbf{u}_i) \boldsymbol{\beta} + \theta(\mathbf{u}_i) + \epsilon_j(\mathbf{u}_i), \quad \epsilon_j(\mathbf{u}_i) \stackrel{iid}{\sim} N(0,\sigma^2)
\end{align*}

**Data Objects:**

- $i \in \{1,\dots,n\}$ indexes unique locations.

- $j \in \{1,\dots,n_i\}$ indexes individuals at each location.

- $Y_j(\mathbf{u}_i)$ denotes the observation of individual $j$ at location $\mathbf{u}_i$.

- $\mathbf{x}_j(\mathbf{u}_i) \in \mathbb{R}^{1 \times p}$, where $p=3$ is the number of predictors (excluding intercept).

## Modeling{.midi}

\begin{align*}
  Y_j(\mathbf{u}_i) &= \alpha + \mathbf{x}_j(\mathbf{u}_i) \boldsymbol{\beta} + \theta(\mathbf{u}_i) + \epsilon_j(\mathbf{u}_i), \quad \epsilon_j(\mathbf{u}_i) \stackrel{iid}{\sim} N(0,\sigma^2)
\end{align*}

**Population Parameters:**

- $\alpha \in \mathbb{R}$ is the intercept.

- $\boldsymbol{\beta} \in \mathbb{R}^p$ is the regression coefficients.

- $\sigma^2 \in \mathbb{R}^+$ is the overall residual error (nugget).

**Location-specific Parameters:**

- $\mathbf{u}_i = (\text{latitude}_i, \text{longitude}_i) \in \mathbb{R}^2$ denotes coordinates of location $i$.

- $\theta(\mathbf{u}_i)$ denotes the spatial intercept at location $\mathbf{u}_i$.


## Location-specific Notation

$$\mathbf{Y}(\mathbf{u}_i) = \alpha \mathbf{1}_{n_i} + \mathbf{X}(\mathbf{u}_i) \boldsymbol{\beta} + \theta(\mathbf{u}_i)\mathbf{1}_{n_i} + \boldsymbol{\epsilon}(\mathbf{u}_i)$$

- $\mathbf{Y}(\mathbf{u}_i) = (Y_1(\mathbf{u}_i),\ldots,Y_{n_i}(\mathbf{u}_i))^\top$

- $\mathbf{X}(\mathbf{u}_i)$ is an $n_i \times p$ dimensional matrix with rows $\mathbf{x}_j(\mathbf{u}_i)$.

- $\boldsymbol{\epsilon}(\mathbf{u}_i) = (\epsilon_i(\mathbf{u}_i),\ldots,\epsilon_{n_i}(\mathbf{u}_i))^\top$, where $\epsilon_j(\mathbf{u}_i) \stackrel{iid}{\sim} N(0,\sigma^2)$.

## Full data notation

$$\mathbf{Y} = \alpha \mathbf{1}_{N} + \mathbf{X} \boldsymbol{\beta} + \mathbf{Z}\boldsymbol{\theta} + \boldsymbol{\epsilon}$$

- $\mathbf{Y} = (\mathbf{Y}(\mathbf{u}_1)^\top,\ldots,\mathbf{Y}(\mathbf{u}_{n})^\top)^\top \in \mathbb{R}^N$, with $N = \sum_{i=1}^n n_i$.

- $\mathbf{X} \in \mathbb{R}^{N \times p}$ that stacks $\mathbf{X}(\mathbf{u}_i)$.

- $\boldsymbol{\theta} = (\theta(\mathbf{u}_1),\ldots,\theta(\mathbf{u}_n))^\top \in \mathbb{R}^n$.

- $\mathbf{Z}$ is $N \times n$ dimensional binary matrix. Each row contains a single 1 in column $i$ that corresponds to the location of $Y_j(\mathbf{u}_i)$.















## Analysis of the full Anemia dataset

Consider the following model:

\begin{align*}
  Y_{ij} = \mathbf{x}_{ij} \boldsymbol{\beta} + \theta(\mathbf{u}_i) + \epsilon_{ij}, \quad \epsilon_{ij} \sim N(0,\sigma^2).
\end{align*}

**Data Objects:**

- $i \in \{1,\dots,n\}$ indexes unique locations.

- $j \in \{1,\dots,n_i\}$ indexes individuals at each location.

- $Y_{ij}$ denotes the hemoglobin level of individual $j$ at location $i$.

- $\mathbf{x}_{ij} \in \mathbb{R}^p$, $p=3$, are covariates including an intercept, age (years), and urban (binary).

## Analysis of the full Anemia dataset

Consider the following model:

\begin{align*}
  Y_{ij} = \mathbf{x}_{ij} \boldsymbol{\beta} + \theta(\mathbf{u}_i) + \epsilon_{ij}, \quad \epsilon_{ij} \sim N(0,\sigma^2).
\end{align*}

**Population Parameters:**

- $\boldsymbol{\beta} \in \mathbb{R}^p$ is the regression coefficients.

- $\sigma^2 \in \mathbb{R}^+$ is the overall residual error (nugget).

## Analysis of the full Anemia dataset

Consider the following model:

\begin{align*}
  Y_{ij} = \mathbf{x}_{ij} \boldsymbol{\beta} + \theta(\mathbf{u}_i) + \epsilon_{ij}, \quad \epsilon_{ij} \sim N(0,\sigma^2).
\end{align*}

**Location-specific Parameters:**

- $\theta(\mathbf{u}_i)$ denotes the spatial intercept at location $\mathbf{u}_i$.

- $\mathbf{u}_i \in \mathbb{R}^d$ denotes the spatial location of location $i$. For example, $\mathbf{u}_i = (\text{latitude}_i, \text{longitude}_i)$, so that $d = 2$.

## Analysis of the full Anemia dataset {.midi}

Rewriting the model in a vectorized form (as a LMM):

$$\mathbf{Y}_i = \mathbf{X}_i \boldsymbol{\beta} + \mathbf{Z}_i \boldsymbol{\theta}(\mathbf{u}_i) + \boldsymbol{\epsilon}_i, \quad \boldsymbol{\epsilon}_i \stackrel{ind}{\sim} N_{n_i}(\mathbf{0}_{n_i}, \sigma^2 \mathbf{I}_{n_i}).$$

-   $\mathbf{Y}_i = (Y_{i1},\ldots,Y_{in_i})$ are location-specific observations.

-   $\boldsymbol{\epsilon}_i = (\epsilon_{i1},\ldots,\epsilon_{in_i})$, such that $\epsilon_{ij} \stackrel{iid}{\sim} N(0,\sigma^2)$.


\begin{align*}
\mathbf{Y}_i = \mathbf{X}_i\boldsymbol{\beta} + \mathbf{Z}_i\boldsymbol{\theta}(\mathbf{u}) + \boldsymbol{\epsilon}, \quad \boldsymbol{\epsilon} \sim N(0,\sigma^2 \mathbf{I}),
\end{align*}

where

- $\mathbf{y} \in \mathbb{R}^{n}$, $X \in \mathbb{R}^{n \times p}$, and $\boldsymbol{\theta}(\mathbf{u}) \in \mathbb{R}^{q}$
- $W \in \mathbb{R}^{n \times q}$ assigns the spatial intercepts to each individual

Consider a Bayesian analysis with 

- $\boldsymbol{\theta}(u) \sim GP(0,K(\psi))$ for some covariance kernel $K$ with parameter $\psi$
- $\boldsymbol{\beta} \sim N(0,\tau^2)$
- and appropriate priors for $\sigma^2$, $\tau^2$, and $\psi$



## Analysis of the full Anemia dataset

Rewriting the model in a vectorized form:
$$
\begin{align}
  \mathbf{y} = X\boldsymbol{\beta} + W\boldsymbol{\theta}(\mathbf{u}) + \boldsymbol{\epsilon}, \quad \boldsymbol{\epsilon} \sim N(0,\sigma^2 \mathbf{I}),
\end{align}
$$
where

- $\mathbf{y} \in \mathbb{R}^{n}$, $X \in \mathbb{R}^{n \times p}$, and $\boldsymbol{\theta}(\mathbf{u}) \in \mathbb{R}^{q}$
- $W \in \mathbb{R}^{n \times q}$ assigns the spatial intercepts to each individual

Consider a Bayesian analysis with 

- $\boldsymbol{\theta}(u) \sim GP(0,K(\psi))$ for some covariance kernel $K$ with parameter $\psi$
- $\boldsymbol{\beta} \sim N(0,\tau^2)$
- and appropriate priors for $\sigma^2$, $\tau^2$, and $\psi$

## Inference goals

Our goal is to [TBU]:

1. study the associations between the outcome and the covariates
2. learn the spatial intercept surface

To achieve these goals, we need to

1. perform Bayesian model fitting to obtain posterior samples of $\boldsymbol{\beta}$, $\boldsymbol{\theta}(\mathbf{u})$ and other parameters
2. make out-of-sample predictions for $\boldsymbol{\theta}(\mathbf{u}^*)$, spatial intercepts at $q^*$ new locations $\mathbf{u}^*=(u_1^*,\dots,u_{q^*})$

## Computational issues with GP

Effectively, the prior for $\boldsymbol{\theta}(\mathbf{u})$ is 
$$
\begin{align}
  \boldsymbol{\theta}(\mathbf{u}) = \begin{pmatrix}
    \theta(u_1) \\
    \vdots \\
    \theta(u_q)
  \end{pmatrix} \sim N(0, \Sigma), \quad \text{where } \Sigma \in \mathbb{R}^{q \times q}, \quad \Sigma_{ij} = K(u_i,u_j|\psi).
\end{align}
$$

This is not scalable because of the need to invert a $q \times q$ covariance matrix for each MCMC iteration, which requires $\mathcal{O}(q^3)$ floating point operations (flops), and $\mathcal{O}(q^2)$ memory.

## Scalable GP methods overview

Existing scalable GP methods broadly fall under two categories.

1. Sparsity methods

  - sparsity in $\Sigma$, e.g., covariance tapering (@furrer2006covariance)
  - sparsity in $\Sigma^{-1}$, e.g., Vecchia approximation (@vecchia1988estimation) and nearest-neighbor GP (@datta2016hierarchical)
  
2. Low-rank methods

  - approximate $\Sigma$ on a low-dimensional subspace
  - e.g., process convolution (@higdon2002space), inducing point method(@snelson2005sparse)


## Hilbert space method for GP

@solin2020hilbert introduced a Hilbert space method for reduced-rank Gaussian process regression (HSGP).

@riutort2023practical discussed how to practically implement HSGP.

Tutorial codes are available in different probabilistic programming languages:

- [stan](https://github.com/gabriuma/basis_functions_approach_to_GP/tree/master/Paper)
- [NumPyro](https://num.pyro.ai/en/0.15.2/examples/hsgp.html)
- [pyMC](https://juanitorduz.github.io/hsgp_intro/)

## Outline

Lecture today:

 - What is HSGP?
 - Theory behind HSGP
 - Why HSGP is scalable?
 - How to use HSGP to make out-of-sample predictions?
 
Lecture on Thursday:

 - Parameter tuning for HSGP
 - How to implement HSGP in `stan`
 - Codes demo for the Anemia dataset
 - Application exercise

## Key idea of HSGP

HSGP approximates $\Sigma$ with

$$\Sigma \approx \Phi(\mathbf{u}) \Lambda(\psi) \Phi(\mathbf{u})^T,$$
where 

- let $m$ denote the number of basis functions
- $\Phi(\mathbf{u}) \in \mathbb{R}^{q \times m}$ only depends on the observed locations
- $\Lambda(\psi) \in \mathbb{R}^{m \times m}$ is diagonal

Computation cost for the GP in each MCMC iteration reduces to $\mathcal{O}(qm + m)$.


# Theory behind HSGP


## Definitions

Let $\mathcal{U} \in \mathbb{R}^d$, $d=2$ be the geographical area of interest. For any locations $u,u' \in \mathcal{U}$, let $r=u-u'$. A covariance function $K$ is

- **stationary** (translation invariant) if 

$$K(u,u') = K(r)$$

- **isotropic** (rotation invariant) if 

$$K(u,u')=K(\|r\|)$$

## Spectral density

By the Bochner's theorem, a bounded stationary covariance function $K$ can be represented as

$$K(r) = \frac 1{(2\pi)^d} \int \exp(i w^Tr) \mu(dw)$$

for some positive finite measure $\mu$. If $\mu$ has a density $S(w)$ with respect to the Lebesgue measure, it is called the **spectral density**. 

By the Wiener-Khintchin theorem, $K(r)$ and $S(w)$ are Fourier duals:
$$
\begin{align}
  K(r) &= \frac 1{(2\pi)^d} \int \exp(iw^Tr) S(w) dw \\
  S(w) &= \int K(r) \exp(-i w^Tr) dr.
\end{align}
$$

## Polynomial expansion of the spectral density

For an isotropic covariance function $K(\|r\|)$, the corresponding spectral density $S(w)=S(\|w\|)$. 

For regular covariance functions, $S$ admits a closed form polynomial expansion of $\|w\|^2$:

$$S(\|w\|) = a_0 + a_1(\|w\|^2) + a_2(\|w\|^2)^2 + \dots \tag{1}$$


Note that $-\|w\|^2$ is the transfer function for the Laplace operator $\nabla^2$. Let $\mathcal{F}$ denote the Fourier transform operator, $f$ denote a twice-differentiable real-valued function,

$$[\mathcal{F}(\nabla^2 f)](w) = -\|w\|^2 [\mathcal{F}(f)](w).$$

## Connection with the Laplacian

Hence applying inverse Fourier transform on both sides of equation (1), we have

$$\mathcal{K} = a_0 + a_1(-\nabla^2) + a_2(-\nabla^2)^2 + \dots \tag{2}$$

where $\mathcal{K}$ is the covariance operator such that 

$$\mathcal{K}f(u) = \int K(u,u')f(u')du'.$$

Equation (2) shows the connection between $\nabla^2$ and $\mathcal{K}$. If we are able to approximate $\nabla^2$, we can approximate $\mathcal{K}$.

## Eigenfunction expansion of the Laplacian

On a compact set $\Omega \subset \mathbb{R}^d$, $-\nabla^2$ has a discrete spectrum. Consider the eigenvalue problem for $-\nabla^2$ with Dirichlet boundary conditions:
$$
\begin{align}
  \begin{cases}
    -\nabla^2 \phi_j(x) = \lambda_j \phi_j(x), & x \in \Omega \\
    \phi_j(x) = 0, & x \in \partial \Omega
  \end{cases}
\end{align}
$$
With sufficiently smooth boundary $\partial \Omega$, eigenvalues and eigenfunctions exist. And because $-\nabla^2$ is positive definite Hermitian,

- all the eigenvalues $\lambda_j$'s are real and positive
- the eigenfunctions $\phi_j$'s are orthonormal

$$\int_\Omega \phi_i(u) \phi_j(u) du = 1_{(i=j)}.$$

## Eigenfunction expansion of the Laplacian

Therefore on $\Omega$, for sufficiently smooth function $f$,

$$-\nabla^2 f(u) = \int L(u,u')f(u')du',$$
where the kernel
$$L(u,u') = \sum_{j=1}^\infty \lambda_j \phi_j(u) \phi_j(u').$$

By orthonormality,

$$L^s(u,u')=\sum_{j=1}^\infty \lambda_j^s \phi_j(u) \phi_j(u').$$

## Low rank approximation

Therefore on $\Omega$ under the boundary conditions, 

$$
\begin{align}
  \mathcal{K}f(u) &= [a_0 + a_1(-\nabla^2) + a_2(-\nabla^2)^2 + \dots]f(u) \\
  &= \int [a_0 + a_1 L(u,u') + a_2L^2(u,u') + \dots]f(u')du'.
\end{align}
$$
Recall $\mathcal{K}f(u)= \int K(u,u') f(u') du'$. Hence

$$
\begin{align}
  K(u,u') &= a_0 + a_1 L(u,u') + a_2L^2(u,u') + \dots \\
  &= \sum_{j=1}^\infty (a_0 + a_1 \lambda_j + a_2 \lambda_j^2 + \dots) \phi_j(u) \phi_j(u') \\
  &= \sum_{j=1}^\infty S(\sqrt{\lambda_j})\phi_j(u) \phi_j(u').
\end{align}
$$

## Low rank approximation

Overall, we can approximate the covariance function with

$$
\begin{align}
  K(u,u') &\approx \sum_{j=1}^\infty S(\sqrt{\lambda_j})\phi_j(u) \phi_j(u') \\
  &\approx \sum_{j=1}^m S(\sqrt{\lambda_j})\phi_j(u) \phi_j(u').
\end{align}
$$

- the spectral density $S$ depends on the covariance function $K$
- the eigenvalue $\lambda_j$'s and eigenfunction $\phi_j$'s only depend on the chosen compact domain $\Omega$ and are independent of $K$
- even with an infinite sum, this is still an approximation as the eigenexpansion is restricted to $\Omega$.

## Key takeaways

HSGP approximates the GP covariance function via an eigenfunction expansion of the Laplace operator in a compact set $\Omega$. 

HSGP **only** works:

1. for covariance functions which admits a power spectral density, e.g., the Matern family.
2. under a user-specified compact domain $\Omega$.
3. under a user-specified number of basis functions $m$. The approximation can be made arbitrarily accurate as $m$ and $\Omega$ increase.
4. for $d \le 3$, at most $4$, because given an accuracy level, $m$ scales exponentially in $d$.

## HSGP Summary

Recall, we want to model a spatial intercept
$$\boldsymbol{\theta}(\mathbf{u}) \sim N(0,\Sigma), \quad \Sigma \in \mathbb{R}^{q \times q}, \quad \Sigma_{ij}=K(u_i,u_j|\psi).$$

For isotropic and *nice* covariance function $K$, we can use HSGP to approximate $\Sigma$ as

$$
\begin{align}
  &\Sigma_{ik} \approx \sum_{j=1}^m S_{K,\psi}(\sqrt{\lambda_j})\phi_j(u_i) \phi_j(u_k) \\
  \implies & \Sigma \approx \Phi(\mathbf{u})\Lambda(\psi)\Phi(\mathbf{u})^T, \quad \text{where}
\end{align}
$$

- $\Lambda(\psi) =$ diag$(S_{K,\psi}(\sqrt{\lambda_1}), \dots, S_{K,\psi}(\sqrt{\lambda_m}))$
- $\Phi(\mathbf{u})$ is a $q \times m$ matrix with $\Phi(\mathbf{u})_{ij} = \phi_j(\mathbf{u}_i)$

## Why HSGP is scalable

The HSGP approximation
$$\Sigma \approx \Phi(\mathbf{u})\Lambda(\psi)\Phi(\mathbf{u})^T$$

is scalable because 

- no matrix inversion
- each MCMC iteration requires $\mathcal{O}(qm + m)$ flops, vs $\mathcal{O}(q^3)$ for full GP
- ideally $m \ll q$, but can be faster even for $m>q$

## Model reparameterization

Under HSGP,
$$\boldsymbol{\theta}(\mathbf{u}) \overset{d}{\approx} \Phi(\mathbf{u}) \Lambda(\psi)^{1/2}\mathbf{z}, \quad \text{for } \mathbf{z} \sim N(0,\mathbf{I}).$$

Therefore we can reparameterize the model as 

$$
\begin{align}
  \mathbf{y} &= X\boldsymbol{\beta} + W\boldsymbol{\theta}(\mathbf{u}) + \boldsymbol{\epsilon} \\
  &\approx X\boldsymbol{\beta} + \underbrace{W\Phi(\mathbf{u}) \Lambda(\psi)^{1/2}}_{A(\psi)}\mathbf{z} + \boldsymbol{\epsilon}
\end{align}
$$

Note the resemblance to linear regression:

- $A(\psi) \in \mathbb{R}^{n \times m}$ is a known design matrix given GP parameters $\psi$
- $\mathbf{z}$ is an unknown parameter vector with prior $N(0,\mathbf{I})$

## HSGP model in `stan`

Similarly, we can use the reparameterized model in `stan`. 

`Stan` documentation calls it the [*non-centered parameterization*](https://mc-stan.org/docs/stan-users-guide/efficiency-tuning.html#hierarchical-models-and-the-non-centered-parameterization), and suggests it's usually more computationally efficient.

```{stan output.var = "model_in_stan", eval = FALSE}
transformed data {
  matrix[q,m] PHI = ...;
}
parameters {
  vector[m] z; // standard normal
}
model {
  vector[q] theta = PHI * sqrt_Lambda .* z;
  
  target += normal_lupdf(y | X * beta + W * theta, sigma);
  target += normal_lupdf(z | 0, 1);
  ...
}
```

## Out-of-sample prediction {.small}

We are also interested in $\boldsymbol{\theta}(\mathbf{u}^*)$, the spatial intercepts at $q^*$ new locations $u^*_1, \dots u^*_{q^*}$. Recall under the GP,
$$
\begin{align}
  \begin{pmatrix}
    \boldsymbol{\theta}(\mathbf{u}) \\
    \boldsymbol{\theta}(\mathbf{u}^*)
  \end{pmatrix} \sim N \left(0,
  \begin{pmatrix}
    \Sigma & C \\
    C^T & \Sigma^*
  \end{pmatrix} \right),
\end{align}
$$
where $\Sigma^* \in \mathbb{R}^{q^* \times q^*}$ is the covariance matrix for $\boldsymbol{\theta}(\mathbf{u}^*)$, and $C$ is the cross covariance matrix between $\boldsymbol{\theta}(\mathbf{u})$ and $\boldsymbol{\theta}(\mathbf{u}^*)$.

The conditional distribution is 
$$\boldsymbol{\theta}(\mathbf{u}^*) \mid \boldsymbol{\theta}(\mathbf{u}) \sim N(C^T \Sigma^{-1}\boldsymbol{\theta}(\mathbf{u}), \Sigma^*-C^T\Sigma^{-1}C).$$
For each posterior sample $\boldsymbol{\theta}(\mathbf{u})^{(s)}$, we obtain a posterior predictive sample $\boldsymbol{\theta}(\mathbf{u}^*)^{(s)}$ by drawing from $\boldsymbol{\theta}(\mathbf{u}^*) \mid \boldsymbol{\theta}(\mathbf{u})^{(s)}$.

## Out-of-sample prediction under HSGP {.small}

Under HSGP, approximately
$$
\begin{align}
  \begin{pmatrix}
    \boldsymbol{\theta}(\mathbf{u}) \\
    \boldsymbol{\theta}(\mathbf{u}^*)
  \end{pmatrix} \sim N \left(0,
  \begin{pmatrix}
    \Phi(\mathbf{u})\Lambda(\psi) \Phi(\mathbf{u})^T & \Phi(\mathbf{u})\Lambda(\psi) \Phi(\mathbf{u}^*)^T \\
    \Phi(\mathbf{u}^*)\Lambda(\psi) \Phi(\mathbf{u})^T & \Phi(\mathbf{u}^*)\Lambda(\psi) \Phi(\mathbf{u}^*)^T
  \end{pmatrix} \right).
\end{align}
$$

1. if $m \ge q$, use the same approach
2. if $m < q$ so that $\Phi(\mathbf{u})\Lambda(\psi) \Phi(\mathbf{u})^T$ is not invertible, 
$$\boldsymbol{\theta}(\mathbf{u}^*) \mid \boldsymbol{\theta}(\mathbf{u}) = (\Phi(\mathbf{u}^*)\Lambda(\psi) \Phi(\mathbf{u})^T) \Sigma^+\boldsymbol{\theta}(\mathbf{u})$$
where $\Sigma^+$ is the pseudo inverse of $\Phi(\mathbf{u})\Lambda(\psi) \Phi(\mathbf{u})^T$ such that $\Phi(\mathbf{u})\Lambda(\psi) \Phi(\mathbf{u})^T\Sigma^+ = \mathbf{I}$.


## Out-of-sample prediction under HSGP

Under the reparamterized model,
$$
\begin{align}
  \boldsymbol{\theta}(\mathbf{u}^*) \mid \boldsymbol{\theta}(\mathbf{u}) &= (\Phi(\mathbf{u}^*)\Lambda(\psi) \Phi(\mathbf{u})^{\top})   \Sigma^+\boldsymbol{\theta}(\mathbf{u}) \\
  &= (\Phi(\mathbf{u}^*)\Lambda(\psi) \Phi(\mathbf{u})^T) \Sigma^+\Phi(\mathbf{u}) \Lambda(\psi)^{1/2}\mathbf{z} \\
  &= \Phi(\mathbf{u}^*)\Lambda(\psi)^{1/2}\mathbf{z}.
\end{align}
$$

Therefore for each posterior sample of $\mathbf{z}$ and $\psi$, we can obtain a posterior predictive sample for $\boldsymbol{\theta}(\mathbf{u}^*)$ as
$$\boldsymbol{\theta}(\mathbf{u}^*)^{(s)} = \Phi(\mathbf{u}^*)\Lambda(\psi^{(s)})^{1/2}\mathbf{z}^{(s)}.$$

## HSGP prediction in `stan`

Easy to implement in `stan`.

```{r, echo=TRUE, eval=FALSE}
transformed data {
  matrix[q_pred,m] PHI_pred = ...;
}
generated quantities {
  vector[q_pred] theta_pred = PHI_pred * sqrt_Lambda .* z;
}

```

## Recap

1. Goal: Bayesian spatial analysis on the full Anemia dataset with $n>8,000$ observations at $q \approx 500$ locations
2. Encountered computational issues with GP, thus need to explore scalable GP methods
3. HSGP is a low-rank approximation method based on eigenfunction expansion of the Laplace operator in a compact domain $\Omega$
4. HSGP can be implemented in `stan` for both model fitting and out-of-sample prediction

## Next lecture

1. How to choose the compact domain $\Omega$?
2. How to choose number of basis function $m$?
3. Which covariance functions can we use HSGP for?
4. How to implement HSGP?

## References
::: {#refs}
:::
