---
title: "AE 08: Disease Mapping"
subtitle: "Modeling Standardized Mortality Ratio across North Carolina"
date: "March 20, 2025"
---

::: callout-important
## Due date

Application exercises (AEs) are submitted by pushing your work to the relevant GitHub repo. AEs from Tuesday lectures should be submitted by Friday by 11:59pm ET, and AEs from Thursday lectures should be submitted by Sunday at 11:59pm ET. Because AEs are intended for in-class activities, there are no extensions given on AEs.

-   Final `.qmd` and `.pdf` files pushed to your GitHub repo
-   **Note:** For homeworks and exams, you will also be required to submit your final `.pdf` file submitted on Gradescope
:::

# Getting Started

## Clone the repo & start new RStudio project

-   Go to the course organization at [github.com/biostat725-sp25](https://github.com/biostat725-sp25) organization on GitHub.
-   Click on the repo with the prefix **ae-08-**. It contains the starter documents you need to complete the AE.
-   Click on the green **CODE** button, select **Use SSH** (this might already be selected by default, and if it is, you'll see the text **Clone with SSH**). Click on the clipboard icon to copy the repo URL.
    -   See the [HW 00 instructions](https://biostat725-sp25.netlify.app/hw/hw-00#connect-rstudio-and-github) if you have not set up the SSH key or configured git.
-   In RStudio, go to *File* $\rightarrow$ *New Project* $\rightarrow$ *Version Control* $\rightarrow$ *Git*.
-   Copy and paste the URL of your assignment repo into the dialog box *Repository URL*. Again, please make sure to have *SSH* highlighted under *Clone* when you copy the address.
-   Click *Create Project*, and the files from your GitHub repo will be displayed in the *Files* pane in RStudio.
-   Click `ae-08.qmd` to open the template Quarto file. This is where you will write up your code and narrative for the AE.

# R packages

We will begin by loading R packages that we will use in this AE.

```{r}
#| label: load-packages
#| message: false

library(tidyverse)     # data wrangling and visualization
library(knitr)         # format output
library(rstan)         # Stan
library(bayesplot)     # figures for post Stan inference
library(loo)           # model comparison
# new packages are below
library(sf)            # functions to work with spatial data
library(rnaturalearth) # maps of DRC
library(geodist)       # conversion from geodisic to miles

```

# Data

## COVID Death 

Today, we will motivate areal spatial data analysis and disease mapping by studying deaths attributed to COVID in 2020 across North Carolina at the county level. The data object `covid_nc_2020` is an `sf` object.

-   Variables are:

    -   `name`: county name.

    -   `population`: 2020 population.

    -   `obs_deaths`: observed number of COVID-related deaths in 2020.

    -   `est_deaths`: estimated number of COVID-related deaths in 2020.

    -   `smr`: standardized mortality ratio.

    -   `age`: precentage of residents over 60 years of age.

    -   `poverty`: percentage of residents below the poverty line.
    
    -   `geometry`: contains centroid and boundary information for each county.

The data set is available in your AE repos and is called `covid_nc_2020.shp`.

```{r, message = FALSE, warning = FALSE, eval = FALSE}
covid_nc_2020 <- st_read("covid_nc_2020.shp")
```

```{r, message = FALSE, warning = FALSE, echo = FALSE}
library(DT)
covid_nc_2020 <- st_read("/Users/sib2/Box Sync/Faculty/Education/biostat725-sp25/course-material/data/covid/covid_nc_2020.shp", quiet = TRUE)
DT::datatable(covid_nc_2020)
```

# Mapping Death

```{r}
#| echo: true
#| message: false
#| warning: false
#| fig-align: "center"
#| fig-width: 10
#| fig-height: 5
#| layout-ncol: 1
ggplot(data = covid_nc_2020) +
  geom_sf(fill = "lightblue", color = "black") +
  geom_sf(aes(fill = obs_deaths), shape = 16, size = 2) +
  theme_minimal() +
  labs(title = "2020 COVID Deaths by County",
       fill = "Deaths") +
  scale_fill_viridis_c() +  # Custom labels for the breaks
  coord_sf()  # Ensures the map is properly projected
```

# Exercise

Replicate the analysis above for a different state within DRC (i.e., don't choose Sud-Kivu). Visualize the posterior predictive distribution mean and standard deviation across a grid and compare it to the one from Sud-Kivu.

**Answer:**

```{r}
#| label: ex1

# add code here

```

::: callout-important
To submit the AE:

-   Render the document to produce the PDF with all of your work from today's class.
-   Push all your work to your AE repo on GitHub. You're done! 🎉
:::
